---
title: "Semi-synthetic DA data - Mouse Gastrulation atlas"
output: html_notebook
---

```{r}
library(tidyverse)
library(MouseGastrulationData)
library(SingleCellExperiment)
library(scater)
library(scran)
library(bit64)
library(patchwork)

options(digits=3)
source('~/milo/notebooks/benchmark/benchmark_utils.R')
```

```{r}
figdir <- "~/mount/gdrive/milo/Figures/benchmark_v2/"
if (!dir.exists(figdir)) {
  dir.create(figdir)  
}

```


### Load dataset

Select samples from late time points (even number of replicates)

```{r}
AtlasSampleMetadata %>%
  arrange(stage) %>%
  mutate(sample=factor(sample, levels=unique(sample))) %>%
  ggplot(aes(stage, sample)) +
  geom_point()

late_samples <- AtlasSampleMetadata %>%
  filter(stage %in% c("E7.75", "E8.0", "E8.25", "E8.5")) %>%
  pull(sample)
```

```{r}
embryo_sce <- EmbryoAtlasData(type="processed", samples = late_samples)
```
```{r}
data.frame(colData(embryo_sce)) %>%
  ggplot(aes(as.factor(sample), as.factor(pool))) + geom_point()
```
```{r}
embryo_sce
```


## Preprocessing
```{r}
logcounts(embryo_sce) <- log1p(counts(embryo_sce))
## Exclude zero counts genes
keep.rows <- rowSums(logcounts(embryo_sce)) != 0
embryo_sce <- embryo_sce[keep.rows, ]

dec <- modelGeneVar(embryo_sce)
hvgs <- getTopHVGs(dec, n=5000)

## Drop cells with NAs in corrected pca (low quality)
embryo_sce <- embryo_sce[,apply(reducedDim(embryo_sce, "pca.corrected"), 1, function(x) all(!is.na(x)))]

## Run UMAP (the data has been subsetted)
# embryo_sce <- scater::runPCA(embryo_sce, subset_row=hvgs)
embryo_sce <- scater::runUMAP(embryo_sce, dimred="pca.corrected")
```

```{r, fig.height=6, fig.width=10}
plotReducedDim(embryo_sce, "UMAP", colour_by="celltype", text_by="celltype", point_size=0.1) +
  scale_color_manual(values=EmbryoCelltypeColours) +
  guides(color=guide_legend(override.aes = list(size=2)))
```

Save object 4 benchmark

```{r}
saveRDS(embryo_sce, "/nfs/team205/ed6/data/milo_benchmark/embryo_data_bm.RDS")

embryo_sce <- readRDS("/nfs/team205/ed6/data/milo_benchmark/embryo_data_bm.RDS")
```

Save list of celltype names and sample by sizes

```{r}
for (i in 1:3) {
  pop_size_df <- as.data.frame(table(embryo_sce$celltype)) %>%
  rename(pop=Var1, pop_size=Freq) %>%
  mutate(size_bin = cut(pop_size, breaks = 10)) %>%
  group_by(size_bin) %>%
  sample_n(size=1) 
  write(as.character(pop_size_df$pop), glue("/nfs/team205/ed6/data/milo_benchmark/pop_sample_{i}.txt"))
}
```



## Simulate ONE DA region

```{r, fig.height=7, fig.width=7}
.plot_prob <- function(pop, pop_enr){
  embryo_sce <- add_synthetic_labels_pop(embryo_sce, pop = pop, pop_col="celltype", pop_enr = pop_enr) 
  plotUMAP(embryo_sce, colour_by="Condition1_prob", point_size=0.1) + scale_color_viridis_c(limits=c(0.5,1))
}

pl_ls_1 <- lapply(c(0.7, 0.85, 1), function(x) .plot_prob("Paraxial mesoderm", x))
pl_ls_2 <- lapply(c(0.7, 0.85, 1), function(x) .plot_prob("Epiblast", x))
pl_ls_3 <- lapply(c(0.7, 0.85, 1), function(x) .plot_prob("Erythroid1", x))

wrap_plots(c(pl_ls_1, pl_ls_2, pl_ls_3)) +
  plot_layout(ncol=3, nrow=3, guides='collect') +
```

## Try one method at the time (to pick params)

<!-- ```{r} -->
<!-- embryo_sce <- readRDS("/nfs/team205/ed6/data/milo_benchmark/embryo_data_bm.RDS") -->

<!-- # Build KNN graph for smoothing -->
<!-- X_red_dim = reducedDim(embryo_sce, "pca.corrected")[,1:30] -->
<!-- graph = buildKNNGraph(t(X_red_dim), k = 15)   -->

<!-- ## Simulate labels -->
<!-- seed=2022 -->
<!-- embryo_sce <- add_synthetic_labels(embryo_sce,  -->
<!--                                    n_components = 10,   -->
<!--                                    redDim='pca.corrected',  -->
<!--                                    seed=seed,  -->
<!--                                    knn_graph = graph,  -->
<!--                                    n_replicates = 6) -->
<!-- true_labels <- ifelse(embryo_sce$Condition2_prob < 0.4, "NegLFC", ifelse(embryo_sce$Condition2_prob > 0.6, "PosLFC", "NotDA")) -->
<!-- colData(embryo_sce)[["true_labels"]] <- true_labels -->

<!-- embryo_sce <- runUMAP(embryo_sce, dimred="pca.corrected", name = 'UMAP', n_dimred=1:30) -->
<!-- ``` -->


<!-- Divide cells into true affected subpopulations: take KNN adjacency matrix, filter out edges between cells with different label, then do louvain_clustering at high res -->

<!-- ```{r} -->

<!-- embryo_sce <- cluster_synthetic_labels(embryo_sce, graph) -->
<!-- plotReducedDim(embryo_sce, dimred="UMAP", colour_by="true_labels", point_size=0.2)  -->
<!-- plotReducedDim(embryo_sce, dimred="UMAP", colour_by="true_DA_clust",  -->
<!--                text_by="true_DA_clust",  -->
<!--                point_size=0.2)  -->
<!-- ``` -->

### Cydar

```{r}
library(cydar)

## Cydar
cydar_res <- run_cydar(embryo_sce, tol=3, d=30)

plot(cydar_res$DAres$logFC, -log10(cydar_res$DAres$SpatialFDR), xlab='logFC', ylab="-log10(SpatialFDR)")
abline(v=aveLogCPM(1,mean(cydar_res$Cd$totals)))

cydar_out <- cydar2output(cydar_res$Cd, cydar_res$DAres, out_type = "continuous")

hist(cydar_res$DAres$PValue)
```

### Louvain

```{r}
library(broom)

## Louvain clustering
d=30
condition_col = 'synth_labels'
sample_col = 'synth_samples'

X_red_dim = reducedDim(embryo_sce, "pca.corrected")[,1:d]
sce.graph <- buildKNNGraph(t(X_red_dim), k=50)
louvain.clust <- cluster_louvain(sce.graph)
louvain.clust.ids <- membership(louvain.clust)

condition_vec <- colData(embryo_sce)[[condition_col]]
sample_labels <- colData(embryo_sce)[[sample_col]]
clust.df <- data.frame("cell_id"=colnames(embryo_sce), "Louvain.Clust"=as.character(louvain.clust.ids))
clust.df$Sample <- sample_labels
clust.df$Condition <- condition_vec

plotUMAP(embryo_sce, colour_by="louvain_clust", text_by="louvain_clust", point_size=0.1)

## Make design matrix
batch_col = NULL
design_df <- as.tibble(colData(embryo_sce)[c(sample_col, condition_col, batch_col)]) %>%
    distinct() %>%
    column_to_rownames(sample_col)
 if (is.null(batch_col)) {
    design <- formula(paste('~', condition_col, collapse = ' '))  
  } else {
    design <- formula(paste('~', batch_col, "+", condition_col, collapse = ' '))
  }

louvain.model <- model.matrix(~ synth_labels, data=design_df)
louvain.count <- as.matrix(table(clust.df$Louvain.Clust, clust.df$Sample))

df <- data.frame(louvain.count) %>%
  rename(cluster=Var1, sample=Var2) %>%
  mutate(cluster=factor(cluster)) %>%
  mutate(synth_labels=str_remove(sample, "_.+")) %>%
  mutate(synth_labels) %>%
  group_by(cluster) %>%
  do(model=glm(Freq ~ synth_labels, data=.,  family="poisson")) 

res_df <- sapply(df$model, function(x) summary(x)$coefficients[2,]) %>%
  t()
colnames(res_df) <- c("logFC","Std. Error", "z value",    "Pval" )
louvain.res <- cbind(df, res_df) %>%
  mutate(FDR=p.adjust(Pval, method = "BH"))
rownames(louvain.res) <- louvain.res$cluster

clust.df$logFC <- louvain.res[clust.df$Louvain.Clust, 'logFC']
clust.df$FDR <- louvain.res[clust.df$Louvain.Clust, 'FDR']

embryo_sce$louvain_logFC <- clust.df$logFC
embryo_sce$louvain_clust <- clust.df$Louvain.Clust
embryo_sce$louvain_FDR <- -log10(clust.df$FDR)

plot(embryo_sce$louvain_logFC, -log10(embryo_sce$lo))
plotUMAP(embryo_sce, colour_by="Condition1_prob", point_size=0.1)
plotUMAP(embryo_sce, colour_by="louvain_logFC", text_by="louvain_clust", point_size=0.1)

```
```{r}
louv_res <- run_louvain(embryo_sce, condition_col = "synth_labels", sample_col="synth_samples", k=50, reduced.dim = "pca_batch")

embryo_sce$louvain_logFC <- louv_res$logFC
embryo_sce$louvain_clust <- louv_res$Louvain.Clust
embryo_sce$louvain_FDR <- -log10(louv_res$FDR)

plot(embryo_sce$louvain_logFC, embryo_sce$louvain_FDR)
plotUMAP(embryo_sce, colour_by="Condition1_prob", point_size=0.1)
plotUMAP(embryo_sce, colour_by="louvain_logFC", point_size=0.1)
plotUMAP(embryo_sce, colour_by="louvain_FDR", text_by="louvain_clust", point_size=0.1)
```


```{r}
# louvain.dge <- DGEList(counts=louvain.count, lib.size=log(colSums(louvain.count)))
# louvain.dge <- estimateDisp(louvain.dge, louvain.model)
# louvain.fit <- glmQLFit(louvain.dge, louvain.model, robust=TRUE)
# louvain.res <- as.data.frame(topTags(glmQLFTest(louvain.fit, coef=ncol(louvain.model)), sort.by='none', n=Inf))
# 
# plotQLDisp(louvain.fit, cex = 3)
clust.df$logFC <- louvain.res[clust.df$Louvain.Clust, 'logFC']
clust.df$FDR <- louvain.res[clust.df$Louvain.Clust, 'FDR']

embryo_sce$louvain_logFC <- clust.df$logFC
embryo_sce$louvain_clust <- clust.df$Louvain.Clust
embryo_sce$louvain_FDR <- clust.df$FDR

louvain.res

plot(embryo_sce$louvain_logFC, -log10(embryo_sce$lo))
plotUMAP(embryo_sce, colour_by="Condition1_prob", text_by="louvain_clust", point_size=0.1)
plotUMAP(embryo_sce, colour_by="louvain_logFC", text_by="louvain_clust", point_size=0.1)
```



<!-- ```{r} -->
<!-- df_glm <- -->
<!--   reshape2::melt(louvain.count, varnames=c("cluster", "sample")) %>% -->
<!--   mutate(cluster=factor(cluster)) %>% -->
<!--   left_join(design_df %>% rownames_to_column("sample")) %>% -->
<!--   group_by(sample) %>% -->
<!--   mutate(n=sum(value)) %>% -->
<!--   ungroup() %>% -->
<!--   mutate(other=n-value) -->

<!-- formula = cbind(value, other) ~ cluster*synth_labels -->
<!-- # ggplot(df_glm[df_glm$cluster=='18',], aes(x=synth_labels, y=value)) + geom_point() -->
<!-- model1 <- glm(formula = formula, family = binomial(link="logit"), data = df_glm) -->
<!-- summary(model1) -->
<!-- ``` -->

### Milo

Pick K using within/between cluster distance
```{r}
nhoodRatio <- function(x, d=50,  reduced.dim="PCA", ...){
  require(BiocNeighbors)
  # calcuate the ratio of within-nhood distances (distance from index to kth cell)
  # to the between nhood-distance (distance to nearest other index cell)
  if(length(nhoodIndex(x)) > 0){
    # get the maximum distance in each nhood
    within.dist <- sapply(unlist(nhoodIndex(x)), FUN=function(X) max(nhoodDistances(x)[[paste0(X)]]))
    # find the 1NN between nhoodIndex cells
    between.dist <- findKNN(reducedDim(x[, unlist(nhoodIndex(x))], reduced.dim)[, c(1:d)],
                            k=1)$distance
    dist.ratio <- within.dist/between.dist
    return(dist.ratio)
  } else{
    stop("Compute nhoods first")
  }
  
}

calc_nhood_ratio <- function(milo, k=50,d=30, prop=0.1, reduced.dim="pca.corrected"){
  milo <- buildGraph(milo, k=k, d=d, reduced.dim = reduced.dim)
  milo <- makeNhoods(milo, prop = prop, k=k, d=d, reduced_dims = reduced.dim)
  d.ratio <- nhoodRatio(milo, d=30)
  d.ratio
  }

milo <- Milo(embryo_sce)
milo <- buildGraph(milo, k=50, d=30, reduced.dim = "pca.corrected")
milo <- makeNhoods(milo, prop = 0.1, k=50, d=30, reduced_dims = "pca.corrected")
milo <- calcNhoodDistance(milo, d=30, reduced.dim = "pca.corrected")

X_pca <- reducedDim(milo, "pca.corrected")[,1:d]

X_pca_ixs <- X_pca[unlist(nhoodIndex(milo)),]
X_pca_cells <- X_pca[rowSums(nhoods(milo))>0,]

nhoods_mat <- nhoods(milo)
as.matrix(nhoods_mat[1:3000, 1:100]) %*% X_pca_ixs[1:100,]



X_pca[nhoods(milo)[1,]==1,]

d.ratios <- sapply(seq(10,50, by=5), function(k) calc_nhood_ratio(milo, k=k))
```

```{r, fig.height=8, fig.width=8}
milo_out <- run_milo(embryo_sce, condition_col="synth_labels", sample_col="synth_samples", reduced.dim="pca.corrected",
         k=50, prop=0.1)

milo_out$Milo <- buildNhoodGraph(milo_out$Milo)
embryo_sce$milo_res <- milo2output(milo_out$Milo, milo_out$DAres, out_type = "labels")
true_labels <- ifelse(embryo_sce$Condition2_prob < 0.4, "NegLFC", ifelse(embryo_sce$Condition2_prob > 0.6, "PosLFC", "NotDA"))
colData(embryo_sce)[["true_labels"]] <- true_labels

plotReducedDim(embryo_sce, dimred="UMAP", colour_by="milo_res", point_size=0.2) 
plotReducedDim(embryo_sce, dimred="UMAP", colour_by="Condition2_prob", point_size=0.2) 
plotReducedDim(embryo_sce, dimred="UMAP", colour_by="true_labels", point_size=0.2) 
```

Compare with MELD

```{r, fig.height=8, fig.width=8}
meld_out <- run_meld_reticulate(embryo_sce, condition_col="synth_labels", sample_col="synth_samples", reduced.dim="pca.corrected",
         k=50)

embryo_sce$meld_res <- meld2output(meld_out, out_type = "labels")

plotReducedDim(embryo_sce, dimred="UMAP", colour_by="meld_res", point_size=0.2) 
plotReducedDim(embryo_sce, dimred="UMAP", colour_by="true_labels", point_size=0.2) 
```

<!-- Does the prop/k need to reach saturation? -->

<!-- ```{r} -->
<!-- milo_ls <- lapply(seq(20,50, by = 10), function(k){ -->
<!--   m <- buildGraph(Milo(embryo_sce), k=k, d=30, reduced.dim = 'pca.corrected')  -->
<!--   m <-  makeNhoods(m, prop = 0.1, k=k, d=30, reduced_dims = 'pca.corrected')  -->
<!--   m -->
<!--   }) -->

<!-- names(milo_ls) <- seq(20,50, by = 10) -->
<!-- imap(milo_ls, ~ data.frame( n_nhoods = rowSums(.x@nhoods), k=.y) ) %>% -->
<!--   purrr::reduce(bind_rows) %>% -->
<!--   ggplot(aes(fill=k, n_nhoods)) + geom_histogram(bins=50, position="identity") -->

<!-- ``` -->

```{r}
## Make design matrix
sample_col = "synth_samples"
condition_col = "synth_labels"
batch_col = NULL
design_df <- as.tibble(colData(embryo_sce)[c(sample_col, condition_col, batch_col)]) %>%
    distinct() %>%
    column_to_rownames(sample_col)
  if (is.null(batch_col)) {
    design <- formula(paste('~', condition_col, collapse = ' '))  
  } else {
    design <- formula(paste('~', batch_col, "+", condition_col, collapse = ' '))
  }

milo_ls <- lapply(milo_ls, function(m){
  ## Test DA
  m <- countCells(m, meta.data = data.frame(colData(m)), sample='synth_samples')
  m <- calcNhoodDistance(m, d=30, reduced.dim = "pca.corrected")
  DA_results <- testNhoods(m, design = design, design.df = design_df)
  return(list(Milo=m, da_res=DA_results))
})

milo_outcome <- lapply(seq_along(milo_ls), function(i) 
  data.frame(pred=milo2output(milo_ls[[i]]$Milo, milo_ls[[i]]$da_res, out_type="labels"), 
             true=milo_ls$`20`$Milo$true_labels,
             method=names(milo_ls)[i])
  ) %>%
  purrr::reduce(bind_rows) %>%
  calculate_outcome()

milo_outcome %>%
  ggplot(aes(method, Power)) + geom_point()

milo_outcome %>%
  ggplot(aes(method, TPR)) + geom_point()

```


```{r, fig.height=8, fig.width=8}
milo_out2 <- run_milo(embryo_sce, condition_col="synth_labels", sample_col="synth_samples", reduced.dim="pca.corrected",
         k=30, prop=0.1)

milo_out2$Milo <- buildNhoodGraph(milo_out2$Milo)

hist(rowSums(milo_out2$Milo@nhoods), breaks=100)

milo_out2$DAres %>%
  ggplot(aes(logFC, - log10(SpatialFDR))) + geom_point()
plotNhoodSizeHist(milo_out2$Milo) / 
  plotNhoodGraphDA(milo_out2$Milo, milo_out2$DAres)
```

### MELD

```{r}
meld_res_ls <- lapply(seq(20,50, by = 10), function(k){
  meld_res <- run_meld_reticulate(embryo_sce, condition_col="synth_labels", sample_col="synth_samples", reduced.dim = "pca.corrected", d=30, k=k)
  meld_out <- meld2output(meld_res, out_type = "labels")
  meld_out
})

lapply(meld_res_ls)

# meld_outcome <- 
  lapply(seq_along(meld_res_ls), function(i) 
  data.frame(pred=meld_res_ls[[i]], 
             true=embryo_sce$true_labels,
             method=seq(20,50, by = 10)[i])
  ) %>%
    purrr::reduce(bind_rows) %>%
    mutate(pred=as.character(pred)) %>%
    mutate(pred=case_when(pred=="PosLFC" ~ "NegLFC",
                        pred=="NegLFC" ~ "PosLFC",
                        TRUE ~ pred)) %>%
    mutate(outcome=case_when(true==pred & pred!="NotDA" ~ 'TP',
                             true!=pred & pred!="NotDA" ~ 'FP',
                             true!=pred & pred=="NotDA" ~ 'FN',
                             true==pred & pred=="NotDA"  ~ "TN"
    )) %>%
    group_by(method, outcome) %>%
    summarise(n=n()) %>%
    pivot_wider(id_cols=method, names_from=outcome, values_from=n, values_fill=0) %>%
    mutate(TPR=TP/(TP+FP), FPR=FP/(TP+FP), TNR=TN/(TN+FN), FNR = FN/(FN+TP),
           Power = 1 - FNR,
           Accuracy = (TP + TN)/(TP + TN + FP + FN),
           Recall = TP / (TP+FN)
    )
```

```{r}
sce <- embryo_sce[,1:5000]

daseq_res <- run_daseq(sce, condition_col="synth_labels", k.vec = c(10,20,30))

hist(daseq_res$da.ratio)
```


## + Batch effect 
(the following code is wrapped in the `run_batch_benchmark.R` script but I am rerunning for viz)

Visualize batch effects of different intensity

```{r, fig.width=9, fig.height=8}
embryo_sce <- readRDS("/nfs/team205/ed6/data/milo_benchmark/embryo_data_bm.RDS")

seed=2024
embryo_sce <- add_synthetic_labels(embryo_sce, 
                                   n_components = 10,  
                                   redDim='pca.corrected', 
                                   seed=seed, 
                                   n_replicates = 3)
true_labels <- ifelse(embryo_sce$Condition2_prob < 0.4, "NegLFC", ifelse(embryo_sce$Condition2_prob > 0.6, "PosLFC", "NotDA"))
colData(embryo_sce)[["true_labels"]] <- true_labels

## Simulate batch effects of different magnitude
bm_sce_ls <- lapply(c(0, 0.25, 0.75, 1), function(sd){
  sce_be <- add_batch_effect(embryo_sce, norm_sd=sd, batch_col = "synth_batches")
  sce_be$norm_sd <- sd
  sce_be
  })

bm_sce_ls <- lapply(bm_sce_ls, function(x) runUMAP(x, dimred="pca_batch", name = 'umap_batch', n_dimred=1:30))

wrap_plots(lapply(bm_sce_ls, function(x) plotReducedDim(x, dimred="umap_batch", colour_by="synth_batches", point_size=0.2) + 
         ggtitle(paste0("Batch effect SD = ", x$norm_sd[1])))) +
  plot_layout(ncol=2, nrow=2, guides="collect") &
  guides(color=guide_legend(override.aes=list(size=2)))
```

```{r}
lapply(bm_sce_ls, function(x) plotReducedDim(x, dimred="umap_batch", colour_by="true_labels", point_size=0.2) + 
         ggtitle(paste0("Batch effect SD = ", x$norm_sd[1])))
```



### Load results from `run_batch_benchmark.R`

```{r}
.outcome_by_prob <- function(benchmark_df, da_upper){
  DA_thresh <- 1 - da_upper
  benchmark_df <- benchmark_df %>%
  mutate(true = ifelse(true_prob < DA_thresh, "NegLFC", "NotDA")) 
  if (benchmark_df$method[1]=="meld") {
    benchmark_df <- mutate(benchmark_df, pred = ifelse(Condition2 < DA_thresh, "NegLFC", "NotDA")) 
  }
  ## Check if conditions were swapped in test
  pred_cor <- benchmark_df %>%
    mutate(pred=factor(pred, levels=c("NegLFC", "NotDA", "PosLFC"), ordered = TRUE)) %>%
    mutate(pred=as.numeric(pred)) %>%
    summarise(cor(pred, true_prob))
  ## Swap outcomes if so
  if (!is.na(pred_cor)) {
    if (pred_cor < 0) {
      benchmark_df <- mutate(benchmark_df, pred = ifelse(pred=="NegLFC", "PosLFC", ifelse(pred=="PosLFC", "NegLFC", "NotDA")))
        }
    }
  
  calculate_outcome(benchmark_df)
}

plot_prob_outcome <- function(out_prob_df, metric='TPR'){
  out_prob_df %>%
  mutate(FDR = FP/(TP+FP)) %>%
  mutate(TPR=ifelse(is.nan(TPR), 0, TPR),
         FPR=ifelse(is.nan(FPR), 0, FPR),
                  FDR=ifelse(is.nan(FDR), 0, FDR),
         Precision=ifelse(is.nan(Precision), 0, Precision)) %>%
  rename(metric=metric) %>%
  group_by(method, DA_thresh, batchEffect, enr) %>%
  summarise(mean_metric=mean(metric), 
            sd_metric=sd(metric)) %>%
  ggplot(aes(DA_thresh, mean_metric, color=method)) +
  geom_point(size=1, alpha=0.7) +
  geom_ribbon(aes(ymin=mean_metric-sd_metric, ymax=mean_metric+sd_metric, fill=method), alpha=0.2, color=NA) +
  geom_line() +
  facet_grid(batchEffect~enr, labeller="label_both") +
  xlab("probability threshold") +
    ylab(paste("mean", metric)) +
  xlim(0.5,0.9) +
    theme_bw(base_size=16)
  
}

## Method color palette
method_names <- c("milo", "daseq", "meld", "louvain", "milo_batch", "louvain_batch")
method_colors <- RColorBrewer::brewer.pal(length(method_names), name="Dark2")
method_colors <- setNames(method_colors, method_names)
method_labels <- c("Milo", "DAseq", "MELD", "Louvain", "Milo (~ batch + condition)", 'Louvain (~ batch + condition)')

outdir <- "/nfs/team205/ed6/data/milo_benchmark/"
res_files <- list.files(outdir, pattern=".DAresults.+.csv")
res_files_full <- list.files(outdir, pattern=".DAresults.+.csv", full.names = TRUE)
in_files <- list.files("~/data/milo_benchmark/synthetic_data/", pattern="coldata.csv")

## Make data frame w benchmark parameters
out_meta_df <- data.frame(file_id = str_remove_all(res_files, "benchmark_embryo_pop_|.csv")) %>%
  separate(col = file_id, sep = ".DAresults.", into=c("file_id", "method")) %>%
  separate(col = file_id, sep = "_enr", into=c("pop", "file_id")) %>%
  separate(col = file_id, sep = "_", into=c("enr", "seed", "batchEffect")) %>%
  # mutate(MNNcorrected=ifelse('MNNcorrected' %in% batchEffect, TRUE, FALSE)) %>%
  mutate(seed=str_remove(seed, "seed"), batchEffect=str_remove_all(batchEffect, "batchEffect|.MNNcorrected")) 

out_meta_df %>%
  filter(method=="louvain_batch" & MNNcorrected) %>%
  ggplot(aes(pop, fill=batchEffect)) + geom_bar() + coord_flip()
```


```{r, warning=FALSE, message=FALSE}
## Read DA predictions
prob_thresh_vec <- seq(0.5, 0.9, 0.05)

outcome_df <- lapply(seq_along(res_files_full), function(i){ 
  print(paste("Outcome no. ", i))
  benchmark_df <- read_csv(res_files_full[i])
  pop_enr <- as.numeric(out_meta_df[i,"enr"])
  lapply(prob_thresh_vec[prob_thresh_vec < pop_enr], function(x){
    benchmark_df %>%
      .outcome_by_prob(da_upper = x) %>%
      mutate(DA_thresh=x) %>%
      ungroup() %>%
      dplyr::select(- method) %>%
      bind_cols(out_meta_df[i,]) %>%
      filter(DA_thresh < as.numeric(enr))
    }) %>%
    purrr::reduce(bind_rows)
  }) %>%
  purrr::reduce(bind_rows)

write_csv(outcome_df, "/nfs/team205/ed6/data/milo_benchmark/outcome_full.csv")
```

```{r}
outcome_df <- read_csv("/nfs/team205/ed6/data/milo_benchmark/outcome_full.csv")
```

Motivate threshold
```{r, warning=FALSE, message=FALSE, fig.width=10, fig.height=4}
metric="TPR"
pl_thr_1 <- outcome_df %>%
  filter(batchEffect==0 & method!="milo_batch") %>%
  mutate(FDR = FP/(TP+FP)) %>%
  mutate(TPR=ifelse(is.nan(TPR), 0, TPR),
         FPR=ifelse(is.nan(FPR), 0, FPR),
                  FDR=ifelse(is.nan(FDR), 0, FDR),
         Precision=ifelse(is.nan(Precision), 0, Precision)) %>%
  rename(metric=metric) %>%
  group_by(method, DA_thresh, batchEffect) %>%
  summarise(mean_metric=mean(metric), 
            sd_metric=sd(metric)) %>%
  ggplot(aes(DA_thresh, mean_metric, color=method)) +
  geom_point(size=2, alpha=0.7) +
  geom_ribbon(aes(ymin=mean_metric-sd_metric, ymax=mean_metric+sd_metric, fill=method), alpha=0.2, color=NA) +
  geom_line() +
  geom_vline(linetype=2, xintercept = 0.6, alpha=0.8) +
  # geom_smooth() +
  scale_color_manual(values = method_colors, labels=setNames(method_labels, method_names)) +
  scale_fill_manual(values = method_colors, labels=setNames(method_labels, method_names)) +
  # facet_grid(batchEffect~., labeller="label_both") +
  xlab("probability threshold") +
    ylab(paste("mean", metric)) +
  xlim(0.5,0.9) +
    theme_bw(base_size=16) 

metric="FDR"
pl_thr_2 <- outcome_df %>%
  filter(batchEffect==0 & method!="milo_batch") %>%
  mutate(FDR = FP/(TP+FP)) %>%
  mutate(TPR=ifelse(is.nan(TPR), 0, TPR),
         FPR=ifelse(is.nan(FPR), 0, FPR),
                  FDR=ifelse(is.nan(FDR), 0, FDR),
         Precision=ifelse(is.nan(Precision), 0, Precision)) %>%
  rename(metric=metric) %>%
  group_by(method, DA_thresh, batchEffect) %>%
  summarise(mean_metric=mean(metric), 
            sd_metric=sd(metric)) %>%
  ggplot(aes(DA_thresh, mean_metric, color=method)) +
  geom_point(size=2, alpha=0.7) +
  geom_ribbon(aes(ymin=mean_metric-sd_metric, ymax=mean_metric+sd_metric, fill=method), alpha=0.2, color=NA) +
  geom_line() +
  geom_vline(linetype=2, xintercept = 0.6, alpha=0.8) +
  scale_color_manual(values = method_colors, labels=setNames(method_labels, method_names)) +
  scale_fill_manual(values = method_colors, labels=setNames(method_labels, method_names)) +
  # facet_grid(batchEffect~., labeller="label_both") +
  xlab("probability threshold") +
    ylab(paste("mean", metric)) +
  xlim(0.5,0.9) +
    theme_bw(base_size=16) 

pl_thr_1 + pl_thr_2 + plot_layout(guides="collect") +
  ggsave(paste0(figdir, 'bm_prob_thresh.pdf'), height = 3, width = 8)
```
```{r}
outcome_df <- outcome_df %>%
  filter(DA_thresh==0.6) 

pl_df <- outcome_df %>%
  filter(method!="milo_batch") %>%
  mutate(FDR = FP/(TP+FP)) %>%
  mutate(TPR=ifelse(is.nan(TPR), 0, TPR),
         FPR=ifelse(is.nan(FPR), 0, FPR),
         FDR=ifelse(is.nan(FDR), 0, FDR),
         Precision=ifelse(is.nan(Precision), 0, Precision)) %>%
  mutate(pop=str_replace(pop, "_", " ")) %>%
  mutate(pop=factor(pop, levels=unique(pop))) %>%
  filter(DA_thresh >= 0.6 & batchEffect==0) 

fig <-  pl_df %>%
  ggplot(aes(method, TPR, color=method)) + 
  geom_boxplot() +
  ggbeeswarm::geom_quasirandom(alpha=0.5) +
  geom_hline(yintercept = 0.8, linetype=2) +
  theme_bw(base_size = 16) +
  scale_color_manual(values = method_colors, labels=setNames(method_labels, method_names)) +
  scale_fill_manual(values = method_colors, labels=setNames(method_labels, method_names)) +
  facet_grid(enr~., labeller = 'label_both') +
pl_df %>%
  ggplot(aes(method, FDR, color=method)) + 
  geom_boxplot() +
  ggbeeswarm::geom_quasirandom(alpha=0.5) +
  geom_hline(yintercept = 0.1, linetype=2) +
  theme_bw(base_size = 16) +
  scale_color_manual(values = method_colors, labels=setNames(method_labels, method_names)) +
  scale_fill_manual(values = method_colors, labels=setNames(method_labels, method_names)) +
  facet_grid(enr~., labeller = 'label_both') +
  plot_layout(guides='collect') &
  theme(axis.text.x=element_blank()) 
fig + ggsave(paste0(figdir, 'bm_nobatch.png'), height = 5, width = 8)
```


Breakdown by population: explain variability in MELD, population size

```{r, fig.width=8, fig.height=7, message=FALSE}
pop_sizes <- table(embryo_sce$celltype)

outcome_df %>%
  filter(method!="milo_batch") %>%
  mutate(FDR = FP/(TP+FP)) %>%
  mutate(TPR=ifelse(is.nan(TPR), 0, TPR),
         FPR=ifelse(is.nan(FPR), 0, FPR),
         FDR=ifelse(is.nan(FDR), 0, FDR),
         Precision=ifelse(is.nan(Precision), 0, Precision)) %>%
  mutate(pop=str_replace(pop, "_", " ")) %>%
  mutate(pop_size=pop_sizes[pop])  %>%
  arrange(pop_size) %>%
  mutate(pop=factor(pop, levels=unique(pop))) %>%
  filter(DA_thresh >= 0.6 & batchEffect==0) %>%
  ggplot(aes(pop_size, TPR, color=method)) + 
  geom_point(size=2, alpha=0.7) +
  facet_grid(enr~., labeller="label_both") +
  geom_smooth(method="lm") +
  theme_bw(base_size = 16) +
  scale_color_manual(values = method_colors, labels=setNames(method_labels, method_names)) +
  scale_fill_manual(values = method_colors, labels=setNames(method_labels, method_names)) +
  xlab("Size of DA population") +
  xlim(0,7000) +
  ggpubr::stat_cor(label.x.npc = 1, label.y.npc = 1, hjust = 1) +
  ggsave(paste0(figdir, "pop_size.png"), height = 7, width = 8)



```


## Batch effect

Load louvain_batch
```{r}
res_files <- list.files(outdir, pattern="batchEffect.{1,4}.DAresults.louvain_batch.csv")
res_files_full <- list.files(outdir, pattern="batchEffect.{1,4}.DAresults.louvain_batch.csv", full.names = TRUE)
in_files <- list.files("~/data/milo_benchmark/synthetic_data/", pattern="coldata.csv")

## Make data frame w benchmark parameters
louvain_out_meta_df <- data.frame(file_id = str_remove_all(res_files, "benchmark_embryo_pop_|.csv")) %>%
  separate(col = file_id, sep = ".DAresults.", into=c("file_id", "method")) %>%
  separate(col = file_id, sep = "_enr", into=c("pop", "file_id")) %>%
  separate(col = file_id, sep = "_", into=c("enr", "seed", "batchEffect")) %>%
  # mutate(MNNcorrected=ifelse('MNNcorrected' %in% batchEffect, TRUE, FALSE)) %>%
  mutate(seed=str_remove(seed, "seed"), batchEffect=str_remove_all(batchEffect, "batchEffect|.MNNcorrected")) 

# louvain_out_meta_df %>%
#   filter(method=="louvain_batch") %>%
#   ggplot(aes(pop, fill=batchEffect)) + geom_bar() + coord_flip()

## Load results
louvainb_outcome_df <- lapply(seq_along(res_files_full), function(i){ 
  print(paste("Outcome no. ", i))
  benchmark_df <- read_csv(res_files_full[i]) %>%
      .outcome_by_prob(da_upper = 0.6) %>%
      mutate(DA_thresh=0.6) %>%
      ungroup() %>%
      dplyr::select(- method) %>%
      bind_cols(louvain_out_meta_df [i,]) 
  pop_enr <- as.numeric(louvain_out_meta_df [i,"enr"])
  benchmark_df
  }) %>%
  purrr::reduce(bind_rows) %>%
  filter(enr > 0.6) 

louvainb_outcome_df <- louvainb_outcome_df %>%
  mutate(seed=as.numeric(seed), batchEffect=as.numeric(batchEffect), enr=as.numeric(enr))
```


<!-- ```{r, fig.width=12, fig.height=10} -->
<!-- data_id = "embryo" -->
<!-- pop="Somitic_mesoderm" -->
<!-- pop_enr=0.8 -->
<!-- seed=43 -->
<!-- outdir <- '/nfs/team205/ed6/data/milo_benchmark/synthetic_data/' -->
<!-- outprefix <- str_c("benchmark_", data_id, "_pop_", pop, '_enr', pop_enr, "_seed", seed) -->
<!-- coldata <- read_csv(paste0(outdir, outprefix, ".coldata.csv")) %>% column_to_rownames() -->
<!-- X_pca <-read_csv(str_c(outdir, outprefix, "_batchEffect", be_sd, ".pca.csv")) %>% column_to_rownames()   -->
<!-- ## Find DA probability x cell -->
<!-- bm_params = list( -->
<!--   milo = list(k=k), -->
<!--   milo_batch = list(k=k), -->
<!--   meld = list(k=k), -->
<!--   daseq = list(k.vec=seq(k-(k/2), k+(k/2), by=10)), -->
<!--   louvain = list(k=k), -->
<!--   louvain_batch = list(k=k) -->
<!--   ) -->

<!-- ## Run DA method -->
<!-- ## Add reduced dim + coldata to sce -->
<!-- colData(embryo_sce) <- DataFrame(coldata) -->
<!-- reducedDim(embryo_sce, "pca_batch") <- as.matrix(X_pca) -->

<!-- ## Run louvain -->
<!-- louvain_batch_res <- run_louvain(embryo_sce, condition_col=condition_col, sample_col=sample_col, -->
<!--                            reduced.dim = "pca_batch", d=30, k=50, batch_col="synth_batches") -->
<!-- louvain_res <- run_louvain(embryo_sce, condition_col=condition_col, sample_col=sample_col, -->
<!--                            reduced.dim = "pca_batch", d=30, k=50) -->

<!-- out <- louvain2output(louvain_batch_res, out_type = 'labels',alpha = 0.1) -->
<!-- lout <- louvain2output(louvain_res, out_type = 'labels', alpha = 0.1) -->

<!-- ggplot(louvain_batch_res, aes(logFC, -log10(FDR))) + -->
<!--   geom_point() + -->
<!--   geom_hline(yintercept = 1) -->

<!-- embryo_sce$louvain <- louvain_batch_res$Louvain.Clust -->
<!-- embryo_sce$pred_louvain_batch <- out -->
<!-- embryo_sce$pred_louvain <- lout -->
<!-- # embryo_sce <- runUMAP(embryo_sce, dimred="pca_batch", name = 'umap_batch', n_dimred=1:30)   -->

<!-- plotReducedDim(embryo_sce, "umap_batch", colour_by="pred_louvain", point_size=0.1) + -->
<!-- plotReducedDim(embryo_sce, "umap_batch", colour_by="pred_louvain_batch", point_size=0.1) + -->
<!-- plotReducedDim(embryo_sce, "umap_batch", colour_by="synth_batches", point_size=0.1) + -->
<!--   plotReducedDim(embryo_sce, "umap_batch", colour_by="synth_labels", point_size=0.1) + -->
<!--   plotReducedDim(embryo_sce, "umap_batch", colour_by="louvain", point_size=0.1) + -->
<!--   plot_layout(nrow=2, ncol=3) -->

<!-- ## Save results -->
<!-- bm <- data.frame(bm_out=out) -->
<!-- bm$true_prob <- embryo_sce$Condition2_prob  -->
<!-- bm$true <- embryo_sce$true_labels -->
<!-- if (!is.null(embryo_sce$true_DA_clust)) { -->
<!-- bm$true_clust <- embryo_sce$true_DA_clust -->
<!-- } -->
<!-- long_bm <- pivot_longer(bm, cols = bm_out, names_to='method', values_to="pred") -->
<!-- .outcome_by_prob(long_bm, 0.6) -->
<!-- long_bm[["method"]] <- method -->

<!-- plotReducedDim(embryo_sce, "umap_batch", colour_by="louvain", point_size=0.1, text_by = "louvain") + -->
<!--   plotReducedDim(embryo_sce, "umap_batch", colour_by="Condition2_prob", point_size=0.1)  -->
<!-- ``` -->

<!-- ```{r} -->
<!-- design_df %>% -->
<!--   ggplot(aes(synth_labels, sample)) + geom_tile() + -->
<!-- design_df %>% -->
<!--   ggplot(aes(synth_batches, sample)) + geom_tile() -->

<!-- ``` -->



<!-- ```{r} -->
<!-- ## Make design matrix -->

<!-- batch_col="synth_batches" -->
<!-- design_df <- as.tibble(colData(embryo_sce)[,c(sample_col, condition_col, batch_col)]) %>% -->
<!--   distinct() %>% -->
<!--   rename(sample=sample_col) -->
<!-- if (is.null(batch_col)) { -->
<!--   design <- formula(paste('Freq ~', condition_col, collapse = ' '))   -->
<!-- } else { -->
<!--   design <- formula(paste('Freq ~', batch_col, "+", condition_col, collapse = ' ')) -->
<!-- } -->
<!-- ## Louvain clustering -->
<!-- sce <- embryo_sce -->
<!-- X_red_dim = reducedDim(sce, reduced.dim)[,1:d] -->
<!-- sce.graph <- buildKNNGraph(t(X_red_dim), k=k) -->
<!-- louvain.clust <- cluster_louvain(sce.graph) -->
<!-- louvain.clust.ids <- membership(louvain.clust) -->

<!-- condition_vec <- colData(sce)[[condition_col]] -->
<!-- sample_labels <- colData(sce)[[sample_col]] -->
<!-- clust.df <- data.frame("cell_id"=colnames(sce), "Louvain.Clust"=as.character(louvain.clust.ids)) -->
<!-- clust.df$Sample <- sample_labels -->
<!-- clust.df$Condition <- condition_vec -->

<!-- louvain.count <- table(clust.df$Louvain.Clust, clust.df$Sample) -->
<!-- attributes(louvain.count)$class <- "matrix" -->

<!-- df <- melt(louvain.count, varnames=c("cluster", "sample"),  value.name="Freq") %>% -->
<!--   # rename(cluster=Var1, sample=Var2) %>% -->
<!--   mutate(cluster=factor(cluster)) %>% -->
<!--   left_join(design_df, by="sample") %>% -->
<!--   group_by(cluster) %>% -->
<!--   do(model=glm.nb(design, data=.))  -->

<!-- summary(df$model[[1]])$coefficients -->
<!-- res_df <- t(sapply(df$model, function(x) summary(x)$coefficients[nrow(summary(x)$coefficients),])) -->
<!-- colnames(res_df) <- c("logFC","Std. Error", "z value",    "Pval" ) -->
<!-- louvain.res <- cbind(df, res_df) %>% -->
<!-- mutate(FDR=p.adjust(Pval, method = "BH")) -->
<!-- rownames(louvain.res) <- louvain.res$cluster -->
<!-- ``` -->

<!-- Test w more replicates -->
<!-- ```{r} -->
<!-- data_id = "embryo" -->
<!-- pop="Somitic mesoderm" -->
<!-- pop_enr=0.8 -->
<!-- seed=43 -->
<!-- outdir <- '/nfs/team205/ed6/data/milo_benchmark/synthetic_data/' -->

<!-- sce <- add_synthetic_labels_pop(sce, pop=pop, pop_column = "celltype", seed=seed, pop_enr=pop_enr, n_replicates = 6, n_batches = 2, n_conditions = 2) -->

<!-- sce_be <- add_batch_effect(sce, batch_col = "synth_batches", norm_sd=0.75) -->

<!-- ## Test DA w louvain -->
<!-- batch_col="synth_batches" -->
<!-- design_df <- as.tibble(colData(sce_be)[,c(sample_col, condition_col, batch_col)]) %>% -->
<!--   distinct() %>% -->
<!--   rename(sample=sample_col) -->
<!-- if (is.null(batch_col)) { -->
<!--   design <- formula(paste('Freq ~', condition_col, collapse = ' '))   -->
<!-- } else { -->
<!--   design <- formula(paste('Freq ~', batch_col, "+", condition_col, collapse = ' ')) -->
<!-- } -->

<!-- ## Louvain clustering -->
<!-- X_red_dim = reducedDim(sce_be, reduced.dim)[,1:d] -->
<!-- sce.graph <- buildKNNGraph(t(X_red_dim), k=k) -->
<!-- louvain.clust <- cluster_louvain(sce.graph) -->
<!-- louvain.clust.ids <- membership(louvain.clust) -->

<!-- condition_vec <- colData(sce_be)[[condition_col]] -->
<!-- sample_labels <- colData(sce_be)[[sample_col]] -->
<!-- clust.df <- data.frame("cell_id"=colnames(sce_be), "Louvain.Clust"=as.character(louvain.clust.ids)) -->
<!-- clust.df$Sample <- sample_labels -->
<!-- clust.df$Condition <- condition_vec -->

<!-- louvain.count <- table(clust.df$Louvain.Clust, clust.df$Sample) -->
<!-- attributes(louvain.count)$class <- "matrix" -->

<!-- df <- melt(louvain.count, varnames=c("cluster", "sample"),  value.name="Freq") %>% -->
<!--   mutate(cluster=factor(cluster)) %>% -->
<!--   left_join(design_df, by="sample") %>% -->
<!--   group_by(sample) %>% -->
<!--   mutate(N_s=sum(Freq)) %>% -->
<!--   ungroup() %>% -->
<!--   group_by(cluster) %>% -->
<!--   do(model=glm(Freq ~ synth_batches + synth_labels, data=., family="poisson"))  -->

<!-- res_df <- t(sapply(df$model, function(x) summary(x)$coefficients[nrow(summary(x)$coefficients),])) -->
<!-- colnames(res_df) <- c("logFC","Std. Error", "z value",    "Pval" ) -->
<!-- louvain.res <- cbind(df, res_df) %>% -->
<!-- mutate(FDR=p.adjust(Pval, method = "BH")) -->
<!-- rownames(louvain.res) <- louvain.res$cluster -->

<!-- sce_be$pred <- louvain_res$FDR < 0.1 -->
<!-- sce_be$louvain <- louvain_res$Louvain.Clust -->

<!-- plotReducedDim(sce_be, "umap_batch", colour_by="pred", point_size=0.1)  -->
<!-- plotReducedDim(sce_be, "umap_batch", colour_by="louvain", text_by="louvain", point_size=0.1)  -->


<!-- ``` -->

<!-- ```{r} -->
<!-- melt(louvain.count, varnames=c("cluster", "sample"),  value.name="Freq") %>% -->
<!--   mutate(cluster=factor(cluster)) %>% -->
<!--     group_by(sample) %>% -->
<!--   mutate(N_s=sum(Freq)) %>% -->
<!--   ungroup() %>% -->
<!--   left_join(design_df, by="sample") %>% -->
<!--   ggplot(aes(synth_batches, Freq/N_s, color=synth_labels)) + -->
<!--   geom_boxplot() + -->
<!--   facet_wrap(cluster~.) -->
<!-- ``` -->


<!-- ```{r} -->
<!-- batch_col=NULL -->
<!-- design_df <- as.tibble(colData(embryo_sce)[,c(sample_col, condition_col, batch_col)]) %>% -->
<!--   distinct() %>% -->
<!--   rename(sample=sample_col) -->
<!-- if (is.null(batch_col)) { -->
<!--   design <- formula(paste('~', condition_col, collapse = ' '))   -->
<!-- } else { -->
<!--   design <- formula(paste('~', batch_col, "+", condition_col, collapse = ' ')) -->
<!-- } -->
<!-- model <- model.matrix(design, data=design_df) -->
<!-- rownames(model) <- rownames(design_df) -->

<!-- subset.counts <- FALSE -->
<!-- keep.nh <- rep(TRUE, nrow(louvain.count)) -->

<!-- dge <- DGEList(counts=louvain.count[keep.nh, ], -->
<!--                    lib.size=log(colSums(louvain.count))) -->

<!-- dge <- estimateDisp(dge, model) -->
<!-- fit <- glmQLFit(dge, model, robust=TRUE) -->
<!-- if(!is.null(model.contrasts)){ -->
<!--     mod.constrast <- makeContrasts(contrasts=model.contrasts, levels=model) -->
<!--     res <- as.data.frame(topTags(glmQLFTest(fit, contrast=mod.constrast), -->
<!--                                  sort.by='none', n=Inf)) -->
<!-- } else{ -->
<!--     n.coef <- ncol(model) -->
<!--     res <- as.data.frame(topTags(glmQLFTest(fit, coef=n.coef), sort.by='none', n=Inf)) -->
<!-- } -->

<!--     res$Nhood <- as.numeric(rownames(res)) -->

<!-- res -->
<!-- ggplot(res, aes(logFC, -log10(FDR))) + geom_point() -->
<!-- ``` -->


<!-- ```{r} -->
<!-- louvainb_outcome_df %>% -->
<!--   filter(enr=='0.8', pop=="Somitic_mesoderm", batchEffect==as.character(be_sd), method=="louvain_batch", seed=='43') -->
<!-- ``` -->





```{r, fig.height=8, fig.width=8}

pl_df <- outcome_df %>%
  bind_rows(louvainb_outcome_df) %>%
  mutate(FDR = FP/(TP+FP)) %>%
  mutate(TPR=ifelse(is.nan(TPR), 0, TPR),
         FPR=ifelse(is.nan(FPR), 0, FPR),
          FDR=ifelse(is.nan(FDR), 0, FDR),
         Precision=ifelse(is.nan(Precision), 0, Precision)) 
metric='TPR'
pl_left<- pl_df %>%
  rename(metric=metric) %>%
  group_by(method, DA_thresh, batchEffect, enr) %>%
  summarise(mean_metric=mean(metric), 
            sd_metric=sd(metric)) %>%
  mutate(group=paste(method, enr)) %>%
  filter(method != "milo" & method != "louvain") %>%
  ggplot(aes(as.numeric(batchEffect), mean_metric, color=method)) +
    geom_line() +
  geom_ribbon(aes(ymin=mean_metric-sd_metric, ymax=mean_metric+sd_metric, fill=method), alpha=0.2, color=NA) +
  geom_point(size=3) +
  facet_grid(enr~., labeller="label_both") +
  xlab("Batch effect magnitude") +
    ylab(paste("mean", metric)) +
    theme_bw(base_size=16) +
    scale_color_manual(values = method_colors, labels=setNames(method_labels, method_names)) +
  scale_fill_manual(values = method_colors, labels=setNames(method_labels, method_names)) +
  geom_hline(yintercept = 0.8, linetype=2) 

metric = "FDR"
pl_right <- pl_df %>%
  rename(metric=metric) %>%
  group_by(method, DA_thresh, batchEffect, enr) %>%
  summarise(mean_metric=mean(metric), 
            sd_metric=sd(metric)) %>%
  mutate(group=paste(method, enr)) %>%
  filter(method != "milo" & method != "louvain") %>%
  ggplot(aes(as.numeric(batchEffect), mean_metric, color=method)) +
  geom_line() +
  geom_ribbon(aes(ymin=mean_metric-sd_metric, ymax=mean_metric+sd_metric, fill=method), alpha=0.2, color=NA) +
  geom_point(size=3) +
  # geom_smooth() +
  facet_grid(enr~., labeller="label_both") +
  xlab("Batch effect magnitude") +
    ylab(paste("mean", metric)) +
    theme_bw(base_size=16) +
  geom_hline(yintercept = 0.1, linetype=2) +
    scale_color_manual(values = method_colors, labels=setNames(method_labels, method_names)) +
  scale_fill_manual(values = method_colors, labels=setNames(method_labels, method_names)) 

pl_left + pl_right + plot_layout(guides="collect") +
  ggsave(paste0(figdir, "batchEffect_linear.png"), height = 7, width = 10) 
```

Visualize outcome in UMAP

```{r, fig.width=16, fig.height=5}
read_input_data <- function(embryo_sce, indir, pop, enr, seed, batchEffect, run_umap=TRUE){
  pca_file <- list.files(indir, pattern=paste0(pop,".+", "enr",enr,".+",'seed', seed, '.+batchEffect',batchEffect,".pca.csv"),
                         full.names = TRUE)
  coldata_file <- list.files(indir, pattern=paste0(pop,".+", "enr",enr,".+",'seed', seed,".coldata.csv"),
                         full.names = TRUE)
  colData(embryo_sce) <- read_csv(coldata_file) %>% column_to_rownames() %>% DataFrame()
  reducedDim(embryo_sce, "pca_batch") <- read_csv(pca_file) %>% column_to_rownames() %>% as.matrix()
  if (run_umap) {
    embryo_sce <- runUMAP(embryo_sce, dimred="pca_batch", name = 'umap_batch', n_dimred=1:30)  
  }
  embryo_sce
}

plot_outcome_umap <- function(sce, method, pop, enr, seed, batchEffect, true=FALSE, data_id="embryo", rasterize=FALSE){
  coldata_file <- list.files(indir, pattern=paste0(".+",data_id,".+",  pop,"_enr",enr,"_seed", seed, ".coldata.csv"),
                           full.names = TRUE)
  colData(sce) <- read_csv(coldata_file) %>% column_to_rownames() %>% DataFrame()
  
  res_file <- list.files(outdir, 
                         pattern= paste0(".+",data_id,".+", pop,"_enr",enr,"_seed", seed, '.+batchEffect',batchEffect,".DAresults.",method,".csv"),
                         full.names = TRUE)
  if (method=="meld") {
    sce$predicted <- ifelse(read_csv(res_file)[["Condition2"]] < 0.3, "NegLFC", 'NotDA')  
  } else if (method=="daseq") {
    sce$predicted <- ifelse(read_csv(res_file)[["pred"]] == "NegLFC", 'PosLFC', ifelse(read_csv(res_file)[["pred"]] == "PosLFC", "NegLFC", "NotDA"))
  } else {
    sce$predicted <- read_csv(res_file)[["pred"]]
  }
  if (true) {
    pl <- data.frame(reducedDim(sce, "umap_batch")) %>%
    rename(UMAP1=X1, UMAP2=X2) %>%
    mutate(true=sce$Condition1_prob) %>%
    arrange(true) %>%
    ggplot(aes(UMAP1, UMAP2, color=true)) + 
    geom_point(size=0.1) +
      scale_color_viridis_c(name="C1 probability", limits=c(0.5,0.9)) +
     theme_classic(base_size=18) +
      guides(color=guide_colorbar(title.position="top", title.hjust = 0.5)) +
     theme(axis.ticks = element_blank(),
           axis.text = element_blank(),
           legend.position = "top") +
     xlab("UMAP1") + ylab("UMAP2") 
  } else {
    pl <- data.frame(reducedDim(sce, "umap_batch")) %>%
    rename(UMAP1=X1, UMAP2=X2) %>%
    mutate(method=method) %>%
    mutate(predicted=factor(sce$predicted, levels=c("NegLFC", "PosLFC", 'NotDA'))) %>%
    arrange(- predicted) %>%
    ggplot(aes(UMAP1, UMAP2, color=predicted)) + 
    geom_point(size=0.1) +
    geom_point(data= . %>% filter(predicted!="NotDA"), size=0.2) +
     scale_color_manual(values=c(NegLFC="red", NotDA="grey", PosLFC="blue"), 
                        labels=c(NegLFC="Enriched in C1", NotDA="No DA", PosLFC="Depleted in C1"), name='') +
     guides(color = guide_legend(override.aes = list(size=1))) +
     theme_classic(base_size=18) +
     theme(legend.position=c(.15,.2), axis.ticks = element_blank(),
           axis.text = element_blank()) +
     xlab("UMAP1") + ylab("UMAP2") +
      facet_wrap(method~.)  
  }
  if (rasterize) {
    pl <- pl + rasterize(geom_point(size=0.1), dpi = 300)
  }
  pl
}

pop="Erythroid2"
enr='0.7'
seed='43'
batchEffect='0'
indir <- "~/data/milo_benchmark/synthetic_data/"

sce <- read_input_data(embryo_sce, indir=indir, pop=pop, enr=enr, seed=seed, batchEffect=batchEffect)

pl_ls <- lapply(c("meld", "milo", "louvain", "daseq"), function(x) plot_outcome_umap(sce, x, pop, enr, seed, batchEffect, true=FALSE))
pl_true <- plot_outcome_umap(sce, 'milo', pop, enr, seed, batchEffect, true=TRUE)
pl_true + 
  wrap_plots(pl_ls, nrow = 1, guides = "collect") +
  plot_layout(widths = c(1,4))
```

Plot benchmark design

```{r, fig.height=4, fig.width=13}
sce <- read_input_data(embryo_sce = embryo_sce, indir=indir, pop="Mesenchyme", enr=0.9, seed=44, batchEffect=0, run_umap = FALSE)
umap_bath_df <- data.frame(reducedDim(sce, "umap_batch")) 

p1 <- umap_bath_df %>%
    rename(UMAP1=X1, UMAP2=X2) %>%
    ggplot(aes(UMAP1, UMAP2)) + 
    geom_point(size=0.1, color="grey") +
    theme_classic(base_size=16) +
      guides(color=guide_colorbar(title.position="top", title.hjust = 0.5)) +
     theme(axis.ticks = element_blank(),
           axis.text = element_blank(),
           legend.position = "top") +
     xlab("UMAP1") + ylab("UMAP2") +
    rasterize(geom_point(size=0.1,  color="grey"), dpi = 300)

p2 <- umap_bath_df %>%
    rename(UMAP1=X1, UMAP2=X2) %>%
    mutate(true=sce$Condition1_prob) %>%
    mutate(true=ifelse(true < 0.5, 0.5, true)) %>%
    arrange(true) %>%
    ggplot(aes(UMAP1, UMAP2, color=true)) + 
    geom_point(size=0.1) +
      scale_color_viridis_c(name="C1 probability") +
    theme_classic(base_size=14) +
      guides(color=guide_colorbar(title.position="left", title.hjust = 1, title.vjust = 1, barwidth = 0.5, barheight = 3)) +
     theme(axis.ticks = element_blank(),
           axis.text = element_blank(),
           legend.position = c(0.75,0.85), legend.background = element_blank()) +
     xlab("UMAP1") + ylab("UMAP2") +
  rasterize(geom_point(size=0.1), dpi = 300)

p3 <- umap_bath_df %>%
    rename(UMAP1=X1, UMAP2=X2) %>%
    mutate(true_labels=sce$synth_labels) %>%
    ggplot(aes(UMAP1, UMAP2, color=true_labels)) + 
    geom_point(size=0.1) +
    scale_color_brewer(palette="Set1", labels=c(Condition1='C1', Condition2="C2"), name="") +
    theme_classic(base_size=16) +
     guides(color=guide_legend(override.aes = list(size=2))) +
     theme(axis.ticks = element_blank(),
           axis.text = element_blank(),
           legend.position = c(0.85,0.95), legend.background = element_blank()) +
     xlab("UMAP1") + ylab("UMAP2") +
  rasterize(geom_point(size=0.1), dpi = 300)

(p1 + p2 + p3) 

```

Design matrix
```{r, fig.height=4, fig.width=13}
des_pl <- data.frame(colData(sce)[,c("synth_labels", "synth_samples")]) %>%
  distinct(synth_labels,synth_samples) %>%
  mutate(synth_labels=str_remove(synth_labels,"ondition"),
         count=1) %>%
  pivot_wider(id_cols = synth_samples, names_from=synth_labels, values_from=count,values_fill=0) %>%
  pivot_longer(cols=c(C1, C2), names_to="synth_labels") %>%
  ggplot(aes(synth_labels, synth_samples)) +
  geom_tile(fill="white", color="black", size=1) +
  geom_tile(data=. %>% filter(value>0), aes(fill=synth_labels), color="black", size=1) +
  theme_classic(base_size=16)   +
  scale_fill_brewer(palette="Set1") +
  guides(fill='none') +
  xlab("Synthetic \ncondition") + ylab('Synthetic samples') +
  theme(axis.text.y = element_blank(),axis.ticks.y = element_blank(),
        axis.line = element_blank())

(p1 + p2 + p3) + des_pl +
  plot_layout(widths=c(1,1,1,0.2), ncol=4) +
  ggsave(paste0(figdir, "bm_example.png"), height = 4, width = 12)
```



```{r, fig.height=8, fig.width=8}
pops <- sample(unique(outcome_df$pop), 3)

p1 <- lapply(pops, function(p) plot_outcome_umap(sce, 'milo', p, 0.7, seed, batchEffect, true=TRUE, rasterize=TRUE))
p2 <- lapply(pops, function(p) plot_outcome_umap(sce, 'milo', p, 0.8, seed, batchEffect, true=TRUE, rasterize=TRUE))
p3 <- lapply(pops, function(p) plot_outcome_umap(sce, 'milo', p, 0.9, seed, batchEffect, true=TRUE, rasterize=TRUE))

fig <- (wrap_plots(p1) /
  wrap_plots(p2) /
  wrap_plots(p3)) +
  plot_layout(guides="collect") &
  theme(legend.position = "right") 

fig +
  ggsave(paste0(figdir, "bm_design_UMAPs.png"), height = 8, width = 10)
```



Plot batch effect

```{r, fig.height=6, fig.width=15}
library(ggrastr)
be_sce_ls <- lapply(c('0',"0.25", "0.5", "0.75", "1"), function(x) read_input_data(embryo_sce, indir, pop, enr, seed, x))

method="milo"

## Plot condition probability 
be_pl_ls <- lapply(be_sce_ls, function(x) plot_outcome_umap(x, method, pop, enr, seed, batchEffect, true=TRUE))

## Plot batch
be_pl_ls2 <- lapply(be_sce_ls, function(sce) {
  coldata_file <- list.files(indir, pattern=paste0(pop,".+", "enr",enr,".+",'seed', seed, ".coldata.csv"),
                           full.names = TRUE)
  colData(sce) <- DataFrame(read_csv(coldata_file) %>% column_to_rownames())
  data.frame(reducedDim(sce, "umap_batch")) %>%
    rename(UMAP1=X1, UMAP2=X2) %>%
    mutate(true=sce$synth_batches) %>%
    ggplot(aes(UMAP1, UMAP2, color=true)) +
    geom_point(size=0.1) +
    scale_color_viridis_d(name="Batch", option="cividis") +
     theme_classic(base_size=18) +
      guides(color=guide_legend(override.aes = list(size=2))) +
     theme(axis.ticks = element_blank(),
           axis.text = element_blank()) +
     xlab("UMAP1") + ylab("UMAP2") +
      rasterize(geom_point(size=0.1), dpi = 300)
  }
)
    
ptop <- wrap_plots(be_pl_ls, nrow=1, guides="collect") &
  theme(legend.position = "right") 
pbottom <- wrap_plots(be_pl_ls2, nrow=1, guides="collect")

(ptop / pbottom) +
  ggsave(paste0(figdir, "batchEffect_UMAPs.pdf"), height=6, width = 13)
```

Design matrix w batch effect

```{r, fig.height=4, fig.width=3}
m1 <- data.frame(colData(sce)[,c("synth_labels", "synth_samples", 'synth_batches')]) %>%
  distinct(synth_labels,synth_samples) %>%
  mutate(synth_labels=str_remove(synth_labels,"ondition"),
         count=1) %>%
  pivot_wider(id_cols = synth_samples, names_from=synth_labels, values_from=count,values_fill=0) %>%
  pivot_longer(cols=c(C1, C2), names_to="synth_labels") %>%
  ggplot(aes(synth_labels, synth_samples)) +
  geom_tile(fill="white", color="black", size=1) +
  geom_tile(data=. %>% filter(value>0), aes(fill=synth_labels), color="black", size=1) +
  theme_classic(base_size=16)   +
  scale_fill_brewer(palette="Set1") +
  guides(fill='none') +
  xlab("Synthetic \ncondition") + ylab('Synthetic samples') +
  theme(axis.text.y = element_blank(),axis.ticks.y = element_blank(),
        axis.line = element_blank())

m2 <- data.frame(colData(sce)[,c("synth_labels", "synth_samples", 'synth_batches')]) %>%
  distinct(synth_labels,synth_samples, synth_batches) %>%
  mutate(synth_labels=str_remove(synth_labels,"ondition"),
         count=1) %>%
  pivot_wider(id_cols = synth_samples, names_from=synth_batches, values_from=count,values_fill=0) %>%
  pivot_longer(cols=c(B1, B2), names_to="synth_batches") %>%
  ggplot(aes(synth_batches, synth_samples)) +
  geom_tile(fill="white", color="black", size=1) +
  geom_tile(data=. %>% filter(value>0), aes(fill=synth_batches), color="black", size=1) +
  theme_classic(base_size=16)   +
  scale_fill_viridis_d(option="cividis") +
  guides(fill='none') +
  xlab("Synthetic \nbatch") + ylab('Synthetic samples') +
  theme(axis.text.y = element_blank(),axis.ticks.y = element_blank(), axis.title.y = element_blank(),
        axis.line = element_blank())

(m1 + m2) +
# (p1 + p2 + p3) + des_pl +
  plot_layout(widths=c(1,1), ncol=2) +
  ggsave(paste0(figdir, "bm_design_batch.png"), height = 4, width = 4)
```

<!-- ```{r, message=FALSE, warning=FALSE, fig.height=8, fig.width=14} -->
<!-- outcome_df %>% -->
<!--   mutate(FDR = FP/(TP+FP)) %>% -->
<!--   mutate(TPR=ifelse(is.nan(TPR), 0, TPR), -->
<!--          FPR=ifelse(is.nan(FPR), 0, FPR), -->
<!--          Precision=ifelse(is.nan(Precision), 0, Precision)) %>% -->
<!--   group_by(method, DA_thresh, batchEffect) %>% -->
<!--   summarise(mean_FDR=mean(FDR),  -->
<!--             sd_FDR=sd(FDR)/sqrt(n()), -->
<!--             mean_TPR=mean(TPR),  -->
<!--             sd_TPR=sd(TPR)/sqrt(n())) %>% -->
<!--   ggplot(aes(DA_thresh, batchEffect)) + -->
<!--   geom_tile(aes(fill=mean_TPR)) + -->
<!--   scale_fill_viridis_c() + -->
<!--   facet_wrap(method~., ncol=2) + -->
<!--   theme_bw(base_size=16) + -->

<!-- outcome_df %>% -->
<!--   mutate(FDR = FP/(TP+FP)) %>% -->
<!--   mutate(TPR=ifelse(is.nan(TPR), 0, TPR), -->
<!--          FPR=ifelse(is.nan(FPR), 0, FPR), -->
<!--          FDR=ifelse(is.nan(FDR), 0, FDR), -->
<!--          Precision=ifelse(is.nan(Precision), 0, Precision)) %>% -->
<!--   group_by(method, DA_thresh, batchEffect) %>% -->
<!--   summarise(mean_FDR=mean(FDR),  -->
<!--             sd_FDR=sd(FDR)/sqrt(n()), -->
<!--             mean_TPR=mean(TPR),  -->
<!--             sd_TPR=sd(TPR)/sqrt(n())) %>% -->
<!--   ggplot(aes(DA_thresh, batchEffect)) + -->
<!--   geom_tile(aes(fill=mean_FDR)) + -->
<!--   scale_fill_viridis_c(direction = -1) + -->
<!--   facet_wrap(method~., ncol=2) + -->
<!--   theme_bw(base_size=16) -->
<!-- ``` -->


## Non-linear batch effect
```{r}
add_batch_effect_nonlinear <- function(sce, batch_col="synth_batches", theta_deg=20, dims=1:2){
  ## Simulate rotation of first 2 PCs
  theta=theta_deg*pi/180
  rotM=matrix(0,nrow=2,ncol=2)
  rotM[1,]=c(cos(theta),-sin(theta))
  rotM[2,]=c(sin(theta),cos(theta))
  
  cellids_sample <- split(sce$cell, sce[[batch_col]])
  X_pca <- reducedDim(sce, "pca.corrected")
  X_pca_batch <- X_pca
  ## Apply rotation to one of the batches
  X_pca_batch[cellids_sample[[names(cellids_sample)[1]]],dims] <- t(rotM %*% t(X_pca_batch[cellids_sample[[names(cellids_sample)[1]]],dims]))
  reducedDim(sce, "pca_batch") <- X_pca_batch
  sce  
  }

coldata_file <- list.files(indir, pattern=paste0(pop,".+", "enr",enr,".+",'seed', seed, ".coldata.csv"),
                           full.names = TRUE)
colData(embryo_sce) <- read_csv(coldata_file) %>% column_to_rownames() %>% DataFrame()
  
batch_col="synth_batches"

embryo_sce <- add_batch_effect_nonlinear(embryo_sce, dims=c(3,5), theta_deg = 30)
embryo_sce <- runUMAP(embryo_sce, dimred="pca_batch", name = 'umap_batch', n_dimred=1:30)  

data.frame(reducedDim(embryo_sce, "umap_batch")) %>%
    rename(UMAP1=X1, UMAP2=X2) %>%
    mutate(true=embryo_sce$synth_batches) %>%
    ggplot(aes(UMAP1, UMAP2, color=true)) +
    geom_point(size=0.1) +
    scale_color_viridis_d(name="Batch", option="cividis") +
     theme_classic(base_size=18) +
      guides(color=guide_legend(override.aes = list(size=2))) +
     theme(axis.ticks = element_blank(),
           axis.text = element_blank()) +
     xlab("UMAP1") + ylab("UMAP2")

data.frame(reducedDim(embryo_sce, "UMAP")) %>%
    rename(UMAP1=X1, UMAP2=X2) %>%
    mutate(true=embryo_sce$synth_batches) %>%
    ggplot(aes(UMAP1, UMAP2, color=true)) +
    geom_point(size=0.1) +
    scale_color_viridis_d(name="Batch", option="cividis") +
     theme_classic(base_size=18) +
      guides(color=guide_legend(override.aes = list(size=2))) +
     theme(axis.ticks = element_blank(),
           axis.text = element_blank()) +
     xlab("UMAP1") + ylab("UMAP2")


# be_sce_ls <- lapply(c('0',"0.25", "0.5", "0.75", "1"), function(x) read_input_data(embryo_sce, indir, pop, enr, seed, x, run_umap = FALSE))

```

Make non-linear batch effect for all populations

```{r}
## Simulate batch effects of different magnitude
save_nonlinear_be <- function(sce, pop, pop_enr, seed){
  ## Load coldata and PCA
  outdir <- '/nfs/team205/ed6/data/milo_benchmark/synthetic_data/'
  outprefix <- str_c("benchmark_embryo_pop_", pop, '_enr', pop_enr, "_seed", seed)
  coldata <- read_csv(paste0(outdir, outprefix, ".coldata.csv")) %>% column_to_rownames()
  ## Add reduced dim + coldata to sce
  colData(sce) <- DataFrame(coldata)
  
  set.seed(seed)
  sce_be <- add_batch_effect_nonlinear(embryo_sce, dims=c(3,5), theta_deg = 30)
  X_pca <- reducedDim(sce_be, "pca_batch")
  ## Save reduced dims
  write_csv(as.data.frame(X_pca) %>% rownames_to_column(), str_c(outdir, outprefix, "_batchEffectNonLinear30.pca.csv"))
  }

pops <- unique(out_meta_df$pop)
pop_enrs <- unique(out_meta_df$enr)
seeds <- unique(out_meta_df$seed)

for (pop in pops) {
  for (pop_enr in pop_enrs[1:3]) {
    for (seed in seeds) {
      save_nonlinear_be(embryo_sce, pop, pop_enr, seed)
    }
  }
}

```


Running in `run_DA_nonlinear.R`. 

Read all results

```{r}
outdir <- "/nfs/team205/ed6/data/milo_benchmark/"
res_files <- list.files(outdir, pattern="NonLinear.+DAresults.+.csv")
res_files_full <- list.files(outdir, pattern="NonLinear.+DAresults.+.csv", full.names = TRUE)

## Make data frame w benchmark parameters
nonl_out_meta_df <- data.frame(file_id = str_remove_all(res_files, "benchmark_embryo_pop_|.csv")) %>%
  separate(col = file_id, sep = ".DAresults.", into=c("file_id", "method")) %>%
  separate(col = file_id, sep = "_enr", into=c("pop", "file_id")) %>%
  separate(col = file_id, sep = "_", into=c("enr", "seed", "batchEffect")) %>%
  mutate(seed=str_remove(seed, "seed"), batchEffect=str_remove(batchEffect, "batchEffect"))

## Load results
nonl_outcome_df <- lapply(seq_along(res_files_full), function(i){ 
  print(paste("Outcome no. ", i))
  benchmark_df <- read_csv(res_files_full[i]) %>%
      .outcome_by_prob(da_upper = 0.6) %>%
      mutate(DA_thresh=0.6) %>%
      ungroup() %>%
      dplyr::select(- method) %>%
      bind_cols(nonl_out_meta_df[i,]) 
  pop_enr <- as.numeric(nonl_out_meta_df[i,"enr"])
  benchmark_df
  }) %>%
  purrr::reduce(bind_rows)

nonl_outcome_df 
```
```{r}
pl_df <- nonl_outcome_df %>%
  filter(enr > 0.6) %>%
  mutate(FDR = FP/(TP+FP)) %>%
  mutate(TPR=ifelse(is.nan(TPR), 0, TPR),
         FPR=ifelse(is.nan(FPR), 0, FPR),
         FDR=ifelse(is.nan(FDR), 0, FDR),
         Precision=ifelse(is.nan(Precision), 0, Precision)) %>%
  mutate(pop=str_replace(pop, "_", " ")) 

pl_df %>%
  ggplot(aes(pop, TPR)) + 
  ggbeeswarm::geom_quasirandom(alpha=0.5) +
  geom_hline(yintercept = 0.8, linetype=2) +
  theme_bw(base_size = 16) +
  scale_color_brewer(palette="Dark2") +
  facet_grid(enr~., labeller = 'label_both') +
pl_df %>%
  ggplot(aes(pop, FDR)) + 
  ggbeeswarm::geom_quasirandom(alpha=0.5) +
  geom_hline(yintercept = 0.1, linetype=2) +
  theme_bw(base_size = 16) +
  scale_color_brewer(palette="Dark2") +
  facet_grid(enr~., labeller = 'label_both') +
plot_layout(guides='collect') 
```
```{r}
lin_be <- outcome_df %>%
  filter(method=="milo_batch" & enr %in% c(0.7, 0.8) & DA_thresh==0.6)

bind_rows(
  mutate(lin_be, class="Linear batch effect"),
  mutate(nonl_outcome_df, class="Non Linear\n batch effect") %>%
    filter(enr > 0.6) %>%
    mutate(enr=as.numeric(enr), seed=as.numeric(seed), batchEffect=as.numeric(str_remove(batchEffect, "NonLinear"))) 
  ) %>%
  ggplot(aes(as.factor(batchEffect), TPR, color=method)) +
  geom_boxplot() +
  ggbeeswarm::geom_quasirandom(alpha=0.5) +
  facet_grid(enr~class, scales="free", space="free") +
  scale_color_manual(values=method_colors[[4]], labels=setNames(method_labels, method_names)) +
  theme_bw(base_size=16) +
  xlab("Batch effect magnitude")
```


### Correcting batch effect

```{r}
library(batchelor)
# 
# be_sce <- be_sce_ls[[4]]
# coldata_file <- list.files(indir, pattern=paste0(pop,".+", "enr",enr,".+",'seed', seed, ".coldata.csv"),full.names = TRUE)
# colData(be_sce) <- DataFrame(read_csv(coldata_file) %>% column_to_rownames())

mnn_correct_batch_effect <- function(be_sce, k=50){
  ## Split in two SCE objects
  b1_be_sce <- be_sce[,be_sce$synth_batches=="B1"]
  b2_be_sce <- be_sce[,be_sce$synth_batches=="B2"]
  
  be_mnn <- reducedMNN(reducedDim(b1_be_sce, "pca.corrected"), reducedDim(b2_be_sce, "pca.corrected"), k=k)
  reducedDim(be_sce, "pca.MNN") <- be_mnn$corrected[colnames(be_sce),]
  be_sce
  }

## Simulate batch effects of different magnitude
save_corrected_be <- function(pop, pop_enr, seed, be_sd){
  ## Load coldata and PCA
  outdir <- '/nfs/team205/ed6/data/milo_benchmark/synthetic_data/'
  outprefix <- str_c("benchmark_embryo_pop_", pop, '_enr', pop_enr, "_seed", seed)
  coldata <- read_csv(paste0(outdir, outprefix, ".coldata.csv")) %>% column_to_rownames()
  X_pca <-read_csv(str_c(outdir, outprefix, "_batchEffect", be_sd, ".pca.csv")) %>% column_to_rownames()
  
  ## Add reduced dim + coldata to sce
  colData(embryo_sce) <- DataFrame(coldata)
  reducedDim(embryo_sce, "pca_batch") <- as.matrix(X_pca)
  
  set.seed(seed)
  sce_be <- mnn_correct_batch_effect(embryo_sce)
  X_pca <- reducedDim(sce_be, "pca.MNN")
  ## Save reduced dims
  write_csv(as.data.frame(X_pca) %>% rownames_to_column(), str_c(outdir, outprefix, "_batchEffect", be_sd,".MNNcorrected.pca.csv"))
  }

be_sds <- unique(out_meta_df$batchEffect)

for (pop in pops) {
    for (seed in seeds) {
      for (be_sd in c(0.5, 0.75, 1)) {
      pop_enr = 0.7
      save_corrected_be(pop, pop_enr, seed, be_sd)
      }
    }
  }
```

<!-- ```{r, fig.width=8, fig.height=8} -->
<!-- X_pca_mnn <- read_csv("/nfs/team205/ed6/data/milo_benchmark/synthetic_data/benchmark_embryo_pop_Somitic_mesoderm_enr0.8_seed45_batchEffect1.MNNcorrected.pca.csv") %>% -->
<!--   column_to_rownames() -->

<!-- coldata <- read_csv("/nfs/team205/ed6/data/milo_benchmark/synthetic_data/benchmark_embryo_pop_Somitic_mesoderm_enr0.8_seed45.coldata.csv") %>% column_to_rownames() -->
<!-- colData(embryo_sce) <- coldata %>% DataFrame() -->
<!-- reducedDim(embryo_sce, "pca_batch") <- as.matrix(X_pca_mnn) -->
<!-- embryo_sce <- runUMAP(embryo_sce, dimred="pca_batch", name="umap_batch") -->
<!-- indir <- "~/data/milo_benchmark/synthetic_data/" -->

<!-- milo_res <- run_milo(embryo_sce, condition_col="synth_labels", sample_col="synth_samples", -->
<!--                          reduced.dim = "pca_batch", d=30, k=50) -->
<!-- milo_res$Milo <- buildNhoodGraph(milo_res$Milo) -->

<!-- plotNhoodGraphDA(milo_res$Milo, milo_res$DAres, layout = "umap_batch") -->

<!-- out <- milo2output(milo_res$Milo, milo_res$DAres, out_type = "label") -->
<!-- embryo_sce$pred <- out -->
<!-- plotReducedDim(embryo_sce, dimred = "umap_batch", point_size=0.1, colour_by="Condition1_prob") -->
<!-- plotReducedDim(embryo_sce, dimred = "umap_batch", point_size=0.1, colour_by="pred") -->
<!-- ``` -->

Read all results

```{r}
outdir <- "/nfs/team205/ed6/data/milo_benchmark/"
res_files <- list.files(outdir, pattern=".+MNNcorrected.+DAresults.+.csv")
res_files_full <- list.files(outdir, pattern=".+MNNcorrected.+DAresults.+.csv", full.names = TRUE)
# in_files <- list.files("~/data/milo_benchmark/synthetic_data/", pattern="_enr0.[78].+coldata.csv")

## Make data frame w benchmark parameters
mnn_out_meta_df <- data.frame(file_id = str_remove_all(res_files, "benchmark_embryo_pop_|.csv")) %>%
  separate(col = file_id, sep = ".DAresults.", into=c("file_id", "method")) %>%
  separate(col = file_id, sep = "_enr", into=c("pop", "file_id")) %>%
  separate(col = file_id, sep = "_", into=c("enr", "seed", "batchEffect")) %>%
  mutate(seed=str_remove(seed, "seed"), batchEffect=str_remove(batchEffect, "batchEffect")) 

## Load results
mnn_outcome_df <- lapply(seq_along(res_files_full), function(i){ 
  print(paste("Outcome no. ", i))
  benchmark_df <- read_csv(res_files_full[i]) %>%
      .outcome_by_prob(da_upper = 0.6) %>%
      mutate(DA_thresh=0.6) %>%
      ungroup() %>%
      dplyr::select(- method) %>%
      bind_cols(mnn_out_meta_df[i,]) 
  pop_enr <- as.numeric(mnn_out_meta_df[i,"enr"])
  benchmark_df
  }) %>%
  purrr::reduce(bind_rows)

mnn_outcome_df %>%
  filter(method=="meld")
```

```{r, fig.height=4, fig.width=11}
no_be <- outcome_df %>%
  filter(batchEffect=="0") %>%
  mutate(FDR = FP/(TP+FP)) 

uncorrected_be <- outcome_df %>%
  filter(batchEffect %in% c("0.5", "0.75", '1')) %>%
  mutate(FDR = FP/(TP+FP)) 

pl_df <- mnn_outcome_df %>%
  filter(enr > 0.6) %>%
   mutate(batchEffect = str_remove(batchEffect, ".MNNcorrected")) %>%
  mutate(FDR = FP/(TP+FP)) %>%
  mutate(TPR=ifelse(is.nan(TPR), 0, TPR),
         FPR=ifelse(is.nan(FPR), 0, FPR),
         FDR=ifelse(is.nan(FDR), 0, FDR),
         Precision=ifelse(is.nan(Precision), 0, Precision)) %>%
  mutate(pop=str_replace(pop, "_", " ")) %>%
  mutate(pop_size=pop_sizes[pop])  %>%
  arrange(pop_size) %>%
  mutate(pop=factor(pop, levels=unique(pop))) 

pl_top <- bind_rows(mutate(no_be, class="no Batch Effect"),
  mutate(uncorrected_be, class="Uncorrected"),
  mutate(pl_df, class="MNN corrected")) %>%
  mutate(class=factor(class, levels=c("no Batch Effect", "MNN corrected", "Uncorrected"))) %>%
    mutate(FDR = FP/(TP+FP)) %>%
  mutate(TPR=ifelse(is.nan(TPR), 0, TPR),
         FPR=ifelse(is.nan(FPR), 0, FPR),
         FDR=ifelse(is.nan(FDR), 0, FDR),
         Precision=ifelse(is.nan(Precision), 0, Precision)) %>%
  filter(method != "milo_batch" & method !="louvain_batch") %>%
  ggplot(aes(as.factor(batchEffect), color=method, TPR)) + 
  geom_boxplot() +
  facet_grid(.~class, space="free", scales="free") +
  scale_color_brewer(palette="Dark2") +
  scale_fill_brewer(palette="Dark2") +
  theme_bw(base_size=16) +
  xlab("Batch effect magnitude") +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(), axis.title.x = element_blank())

pl_bottom <- bind_rows(mutate(no_be, class="no Batch Effect"),
  mutate(uncorrected_be, class="Uncorrected"),
  mutate(pl_df, class="MNN corrected")) %>%
  mutate(class=factor(class, levels=c("no Batch Effect", "MNN corrected", "Uncorrected"))) %>%
    mutate(FDR = FP/(TP+FP)) %>%
  mutate(TPR=ifelse(is.nan(TPR), 0, TPR),
         FPR=ifelse(is.nan(FPR), 0, FPR),
         FDR=ifelse(is.nan(FDR), 0, FDR),
         Precision=ifelse(is.nan(Precision), 0, Precision)) %>%
  filter(method != "milo_batch" & method !="louvain_batch") %>%
  ggplot(aes(as.factor(batchEffect), color=method, FDR)) + 
  geom_boxplot() +
  facet_grid(.~class, space="free", scales="free") +
  scale_color_brewer(palette="Dark2") +
  scale_fill_brewer(palette="Dark2") +
  theme_bw(base_size=16) +
  xlab("Batch effect magnitude")

pl_top / pl_bottom +
  plot_layout(guides="collect")
```

```{r}
mnn_outcome_df %>%
  mutate(batchEffect = as.numeric(str_remove(batchEffect, ".MNNcorrected"))) %>%
  mutate(FDR = FP/(TP+FP)) %>%
  mutate(TPR=ifelse(is.nan(TPR), 0, TPR),
         FPR=ifelse(is.nan(FPR), 0, FPR),
          FDR=ifelse(is.nan(FDR), 0, FDR),
         Precision=ifelse(is.nan(Precision), 0, Precision)) %>%
  rename(metric=metric) %>%
  group_by(method, DA_thresh, batchEffect, enr) %>%
  summarise(mean_metric=mean(metric), 
            sd_metric=sd(metric)) %>%
  mutate(group=paste(method, enr)) %>%
  ggplot(aes(batchEffect, mean_metric, color=method)) +
    geom_line() +
  geom_ribbon(aes(ymin=mean_metric-sd_metric, ymax=mean_metric+sd_metric, fill=method), alpha=0.2, color=NA) +
  geom_point(size=3) +
  facet_grid(enr~., labeller="label_both") +
  xlab("Batch effect magnitude") +
    ylab(paste("mean", metric)) +
    theme_bw(base_size=16) +
  scale_color_brewer(palette="Dark2") +
  geom_hline(yintercept = 0.8, linetype=2) +
  scale_fill_brewer(palette="Dark2") 
```


Plot UMAP
```{r, fig.width=6,fig.height=7}
X_pca_mnn <- read_csv("/nfs/team205/ed6/data/milo_benchmark/synthetic_data/benchmark_embryo_pop_Somitic_mesoderm_enr0.8_seed43_batchEffect0.75.MNNcorrected.pca.csv") %>%
  column_to_rownames()

all(rownames(X_pca_mnn) == colnames(embryo_sce))
coldata <- read_csv("/nfs/team205/ed6/data/milo_benchmark/synthetic_data/benchmark_embryo_pop_Somitic_mesoderm_enr0.8_seed43.coldata.csv")
colData(embryo_sce) <- coldata %>% column_to_rownames() %>% DataFrame()
reducedDim(embryo_sce, "pca_batch") <- as.matrix(X_pca_mnn)
embryo_sce <- runUMAP(embryo_sce, dimred="pca_batch", name="umap_batch")

indir <- "~/data/milo_benchmark/synthetic_data/"
res <- read_csv("/nfs/team205/ed6/data/milo_benchmark/benchmark_embryo_pop_Somitic_mesoderm_enr0.8_seed43_batchEffect0.75.DAresults.milo.csv")
embryo_sce$pred <- res$pred
embryo_sce$true <- res$true
plotReducedDim(embryo_sce, "umap_batch", colour_by="Condition1_prob", point_size=0.1)
plotReducedDim(embryo_sce, "umap_batch", colour_by="pred", point_size=0.1)
plotReducedDim(embryo_sce, "umap_batch", colour_by="true", point_size=0.1)

res %>%
  .outcome_by_prob(da_upper = 0.6) %>%
  mutate(FDR = FP/(TP+FP)) 
```

### Read results for different geometries

#### Linear

```{r}
outdir <- "/nfs/team205/ed6/data/milo_benchmark/"
res_files <- list.files(outdir, pattern="linear.+DAresults")
res_files_full <- list.files(outdir, pattern="linear.+DAresults", full.names = TRUE)
in_files <- list.files("~/data/milo_benchmark/synthetic_data/", pattern="linear.+coldata.csv")

## Make data frame w benchmark parameters
linear_out_meta_df <- data.frame(file_id = str_remove_all(res_files, "benchmark_embryo_pop_|.csv")) %>%
  separate(col = file_id, sep = ".DAresults.", into=c("file_id", "method")) %>%
  separate(col = file_id, sep = "_enr", into=c("pop", "file_id")) %>%
  separate(col = file_id, sep = "_", into=c("enr", "seed", "batchEffect")) %>%
  mutate(seed=str_remove(seed, "seed"), batchEffect=str_remove(batchEffect, "batchEffect")) 

## Load results
linear_outcome_df <- lapply(seq_along(res_files_full), function(i){ 
  print(paste("Outcome no. ", i))
  benchmark_df <- read_csv(res_files_full[i]) %>%
      .outcome_by_prob(da_upper = 0.6) %>%
      mutate(DA_thresh=0.6) %>%
      ungroup() %>%
      dplyr::select(- method) %>%
      bind_cols(linear_out_meta_df[i,]) 
  pop_enr <- as.numeric(linear_out_meta_df[i,"enr"])
  benchmark_df
  }) %>%
  purrr::reduce(bind_rows)

linear_outcome_df 
```
```{r}
pl_df <- linear_outcome_df %>%
  filter(method!="milo_batch") %>%
  mutate(FDR = FP/(TP+FP)) %>%
  mutate(TPR=ifelse(is.nan(TPR), 0, TPR),
         FPR=ifelse(is.nan(FPR), 0, FPR),
         FDR=ifelse(is.nan(FDR), 0, FDR),
         Precision=ifelse(is.nan(Precision), 0, Precision)) %>%
  mutate(pop=str_replace(pop, "_", " ")) %>%
  mutate(pop_size=pop_sizes[pop])  %>%
  arrange(pop_size) %>%
  mutate(pop=factor(pop, levels=unique(pop))) %>%
  filter(DA_thresh >= 0.6 & batchEffect==0) 

pl_right <- pl_df %>%
  ggplot(aes(method, TPR, color=method)) + 
  geom_boxplot() +
  ggbeeswarm::geom_quasirandom(alpha=0.5) +
  geom_hline(yintercept = 0.8, linetype=2) +
  theme_bw(base_size = 16) +
  scale_color_brewer(palette="Dark2") +
  facet_grid(enr~., labeller = 'label_both') +
pl_df %>%
  ggplot(aes(method, FDR, color=method)) + 
  geom_boxplot() +
  ggbeeswarm::geom_quasirandom(alpha=0.5) +
  geom_hline(yintercept = 0.1, linetype=2) +
  theme_bw(base_size = 16) +
  scale_color_brewer(palette="Dark2") +
  facet_grid(enr~., labeller = 'label_both') +
plot_layout(guides='collect') &
  theme(axis.text.x=element_blank())

pl_right
```

Plot UMAP
```{r, fig.width=6,fig.height=7}
linear_sce <- readRDS("~/data/milo_benchmark/linear_data_bm.RDS")
indir <- "~/data/milo_benchmark/synthetic_data/"

names(reducedDims(linear_sce))[2] <- "umap_batch"
colnames(reducedDims(linear_sce)[[2]]) <- c("X1", "X2") 

pl_top <- plot_outcome_umap(linear_sce, method = 'milo', pop="M5", enr = 0.9, seed = 43,batchEffect = 0, true = TRUE, data_id="linear") + coord_fixed() + theme(legend.position="right")

plot_linear <- ((pl_top) / (pl_right)) +
  plot_layout(heights = c(1,2))
plot_linear
```



#### Branching

```{r}
outdir <- "/nfs/team205/ed6/data/milo_benchmark/"
res_files <- list.files(outdir, pattern="branching.+DAresults")
res_files_full <- list.files(outdir, pattern="branching.+DAresults", full.names = TRUE)
in_files <- list.files("~/data/milo_benchmark/synthetic_data/", pattern="branching.+coldata.csv")

## Make data frame w benchmark parameters
branch_out_meta_df <- data.frame(file_id = str_remove_all(res_files, "benchmark_embryo_pop_|.csv")) %>%
  separate(col = file_id, sep = ".DAresults.", into=c("file_id", "method")) %>%
  separate(col = file_id, sep = "_enr", into=c("pop", "file_id")) %>%
  separate(col = file_id, sep = "_", into=c("enr", "seed", "batchEffect")) %>%
  mutate(seed=str_remove(seed, "seed"), batchEffect=str_remove(batchEffect, "batchEffect")) 

## Load results
branch_outcome_df <- lapply(seq_along(res_files_full), function(i){ 
  print(paste("Outcome no. ", i))
  benchmark_df <- read_csv(res_files_full[i]) %>%
      .outcome_by_prob(da_upper = 0.6) %>%
      mutate(DA_thresh=0.6) %>%
      ungroup() %>%
      dplyr::select(- method) %>%
      bind_cols(branch_out_meta_df[i,]) 
  pop_enr <- as.numeric(branch_out_meta_df[i,"enr"])
  benchmark_df
  }) %>%
  purrr::reduce(bind_rows)

branch_outcome_df 
  
```
```{r}
pl_df <- branch_outcome_df %>%
  filter(method!="milo_batch") %>%
  mutate(FDR = FP/(TP+FP)) %>%
  mutate(TPR=ifelse(is.nan(TPR), 0, TPR),
         FPR=ifelse(is.nan(FPR), 0, FPR),
         FDR=ifelse(is.nan(FDR), 0, FDR),
         Precision=ifelse(is.nan(Precision), 0, Precision)) %>%
  mutate(pop=str_replace(pop, "_", " ")) %>%
  mutate(pop_size=pop_sizes[pop])  %>%
  arrange(pop_size) %>%
  mutate(pop=factor(pop, levels=unique(pop))) %>%
  filter(DA_thresh >= 0.6 & batchEffect==0) 

pl_right <- pl_df %>%
  ggplot(aes(method, TPR, color=method)) + 
  geom_boxplot() +
  ggbeeswarm::geom_quasirandom(alpha=0.5) +
  geom_hline(yintercept = 0.8, linetype=2) +
  theme_bw(base_size = 16) +
  scale_color_brewer(palette="Dark2") +
  facet_grid(enr~., labeller = 'label_both') +
pl_df %>%
  ggplot(aes(method, FDR, color=method)) + 
  geom_boxplot() +
  ggbeeswarm::geom_quasirandom(alpha=0.5) +
  geom_hline(yintercept = 0.1, linetype=2) +
  theme_bw(base_size = 16) +
  scale_color_brewer(palette="Dark2") +
  facet_grid(enr~., labeller = 'label_both') +
plot_layout(guides='collect') &
  theme(axis.text.x=element_blank())

pl_right
```

Plot UMAP
```{r, fig.width=6,fig.height=7}
branch_sce <- readRDS("~/data/milo_benchmark/branching_data_bm.RDS")
indir <- "~/data/milo_benchmark/synthetic_data/"

names(reducedDims(branch_sce))[2] <- "umap_batch"
colnames(reducedDims(branch_sce)[[2]]) <- c("X1", "X2") 

pl_top <- plot_outcome_umap(branch_sce, method = 'milo', pop="M5", enr = 0.9, seed = 43,batchEffect = 0, true = TRUE, data_id="branching") + coord_fixed() + theme(legend.position="right")

plot_branch <- ((pl_top) / (pl_right)) +
  plot_layout(heights = c(1,2))
plot_branch
```


#### Simple clustering

```{r}
outdir <- "/nfs/team205/ed6/data/milo_benchmark/"
res_files <- list.files(outdir, pattern="cluster.+DAresults")
res_files_full <- list.files(outdir, pattern="cluster.+DAresults", full.names = TRUE)
in_files <- list.files("~/data/milo_benchmark/synthetic_data/", pattern="cluster.+coldata.csv")

## Make data frame w benchmark parameters
cluster_out_meta_df <- data.frame(file_id = str_remove_all(res_files, "benchmark_embryo_pop_|.csv")) %>%
  separate(col = file_id, sep = ".DAresults.", into=c("file_id", "method")) %>%
  separate(col = file_id, sep = "_enr", into=c("pop", "file_id")) %>%
  separate(col = file_id, sep = "_", into=c("enr", "seed", "batchEffect")) %>%
  mutate(seed=str_remove(seed, "seed"), batchEffect=str_remove(batchEffect, "batchEffect")) 

## Load results
cluster_outcome_df <- lapply(seq_along(res_files_full), function(i){ 
  print(paste("Outcome no. ", i))
  benchmark_df <- read_csv(res_files_full[i]) %>%
      .outcome_by_prob(da_upper = 0.6) %>%
      mutate(DA_thresh=0.6) %>%
      ungroup() %>%
      dplyr::select(- method) %>%
      bind_cols(cluster_out_meta_df[i,]) 
  pop_enr <- as.numeric(cluster_out_meta_df[i,"enr"])
  benchmark_df
  }) %>%
  purrr::reduce(bind_rows)

cluster_outcome_df 
  
```
```{r}
pl_df <- cluster_outcome_df %>%
  filter(method!="milo_batch") %>%
  mutate(FDR = FP/(TP+FP)) %>%
  mutate(TPR=ifelse(is.nan(TPR), 0, TPR),
         FPR=ifelse(is.nan(FPR), 0, FPR),
         FDR=ifelse(is.nan(FDR), 0, FDR),
         Precision=ifelse(is.nan(Precision), 0, Precision)) %>%
  mutate(pop=str_replace(pop, "_", " ")) %>%
  mutate(pop_size=pop_sizes[pop])  %>%
  arrange(pop_size) %>%
  mutate(pop=factor(pop, levels=unique(pop))) %>%
  filter(DA_thresh >= 0.6 & batchEffect==0) 

pl_right <- pl_df %>%
  ggplot(aes(method, TPR, color=method)) +
  geom_boxplot() +
  ggbeeswarm::geom_quasirandom(alpha=0.5) +
  geom_hline(yintercept = 0.8, linetype=2) +
  theme_bw(base_size = 16) +
  scale_color_brewer(palette="Dark2") +
  facet_grid(enr~., labeller = 'label_both') +
pl_df %>%
  ggplot(aes(method, FDR, color=method)) + 
  geom_boxplot() +
  ggbeeswarm::geom_quasirandom(alpha=0.5) +
  geom_hline(yintercept = 0.1, linetype=2) +
  theme_bw(base_size = 16) +
  scale_color_brewer(palette="Dark2") +
  facet_grid(enr~., labeller = 'label_both') +
plot_layout(guides='collect') &
  theme(axis.text.x=element_blank())

pl_right
```
Plot UMAP
```{r, fig.width=14,fig.height=8}
cluster_sce <- readRDS("~/data/milo_benchmark/cluster_data_bm.RDS")
indir <- "~/data/milo_benchmark/synthetic_data/"

names(reducedDims(cluster_sce))[2] <- "umap_batch"
colnames(reducedDims(cluster_sce)[[2]]) <- c("X1", "X2") 
pl_top <- plot_outcome_umap(cluster_sce, method = 'milo', pop="3", enr = 0.9, seed = 43,batchEffect = 0, true = TRUE, data_id="cluster") + theme(legend.position="right")

plot_cluster <- ((pl_top) / (pl_right)) +
  plot_layout(heights = c(1,2))
plot_cluster
```
```{r, fig.height=8, fig.width=14}
fig <- ((plot_cluster) & theme(legend.position="none") |
   (plot_linear) & theme(legend.position="none") |
   (plot_branch)) + 
  plot_layout(widths=c(1,1,1))

fig +
  ggsave(paste0(figdir, "simData_figure.png"), width=14, height = 8)
```


## Cydar on simple trajectory

Pick radius parameter for different datasets: For each hypersphere centred on a cell, the radius
required to include a certain number of nearest neighbours is computed.

```{r}
sample_col = "synth_samples"
d=30
sce <- readRDS("/nfs/team205/ed6/data/milo_benchmark/cluster_data_bm.RDS")
sample_ls <- split(1:ncol(sce), sce[[sample_col]])
processed.exprs <- lapply(sample_ls, function(s) reducedDim(sce[,s], "pca.corrected")[,1:d])
cd <- prepareCellData(processed.exprs)
boxplot(neighborDistances(cd, neighbors=50, as.tol=TRUE))
```

```{r}
sample_col = "synth_samples"
d=30
sce <- readRDS("/nfs/team205/ed6/data/milo_benchmark/branching_data_bm.RDS")
sample_ls <- split(1:ncol(sce), sce[[sample_col]])
processed.exprs <- lapply(sample_ls, function(s) reducedDim(sce[,s], "pca.corrected")[,1:d])
cd <- prepareCellData(processed.exprs)
boxplot(neighborDistances(cd, neighbors=50, as.tol=TRUE))
```

```{r}
sample_col = "synth_samples"
d=30
sce <- readRDS("/nfs/team205/ed6/data/milo_benchmark/linear_data_bm.RDS")
sample_ls <- split(1:ncol(sce), sce[[sample_col]])
processed.exprs <- lapply(sample_ls, function(s) reducedDim(sce[,s], "pca.corrected")[,1:d])
cd <- prepareCellData(processed.exprs)
boxplot(neighborDistances(cd, neighbors=50, as.tol=TRUE))
```
```{r}
sample_col = "synth_samples"
d=30
sce <- embryo_sce
sample_ls <- split(1:ncol(sce), sce[[sample_col]])
processed.exprs <- lapply(sample_ls, function(s) reducedDim(sce[,s], "pca.corrected")[,1:d])
cd <- prepareCellData(processed.exprs)
boxplot(neighborDistances(cd, neighbors=50, as.tol=TRUE))
```

```{r}
tol_dataset <- list(cluster=0.8, branching=6, linear=6, embryo=1)
```



```{r}
sce <- readRDS("/nfs/team205/ed6/data/milo_benchmark/linear_data_bm.RDS")
data_id <- "linear"
pop <- "M1"
pop_enr <- 0.9
seed <- '43'
mnn_correct <- "no"
be_sd <- '0'

## Load coldata and PCA
outdir <- '/nfs/team205/ed6/data/milo_benchmark/synthetic_data/'
outprefix <- str_c("benchmark_", data_id, "_pop_", pop, '_enr', pop_enr, "_seed", seed)
coldata <- read_csv(paste0(outdir, outprefix, ".coldata.csv")) %>% column_to_rownames()
if (mnn_correct=="no") {
  X_pca <-read_csv(str_c(outdir, outprefix, "_batchEffect", be_sd, ".pca.csv")) %>% column_to_rownames()  
} else {
  X_pca <-read_csv(str_c(outdir, outprefix, "_batchEffect", be_sd, ".MNNcorrected.pca.csv")) %>% column_to_rownames()
}
## Add reduced dim + coldata to sce
colData(sce) <- DataFrame(coldata)
reducedDim(sce, "pca_batch") <- as.matrix(X_pca)
plotReducedDim(sce, "UMAP",colour_by="synth_labels")
plotReducedDim(sce, "UMAP",colour_by="Condition1_prob")

cydar_res <- runDA(sce, coldata, X_pca, "cydar", out_type = "labels",d = 30,
                    params=list(cydar=list(tol=tol_dataset[[data_id]], downsample=3)))
df <- .outcome_by_prob(cydar_res, 0.6) 

df

```


```{r}
sce <- embryo_sce
data_id <- "embryo"
pop <- "Erythroid2"
pop_enr <- 0.9
seed <- '43'
mnn_correct <- "no"
be_sd <- '0'

## Load coldata and PCA
outdir <- '/nfs/team205/ed6/data/milo_benchmark/synthetic_data/'
outprefix <- str_c("benchmark_", data_id, "_pop_", pop, '_enr', pop_enr, "_seed", seed)
coldata <- read_csv(paste0(outdir, outprefix, ".coldata.csv")) %>% column_to_rownames()
if (mnn_correct=="no") {
  X_pca <-read_csv(str_c(outdir, outprefix, "_batchEffect", be_sd, ".pca.csv")) %>% column_to_rownames()  
} else {
  X_pca <-read_csv(str_c(outdir, outprefix, "_batchEffect", be_sd, ".MNNcorrected.pca.csv")) %>% column_to_rownames()
}
## Add reduced dim + coldata to sce
colData(sce) <- DataFrame(coldata)
reducedDim(sce, "pca_batch") <- as.matrix(X_pca)
plotReducedDim(sce, "UMAP",colour_by="synth_labels")
plotReducedDim(sce, "UMAP",colour_by="Condition1_prob")

cydar_res <- runDA(sce, coldata, X_pca, "cydar", out_type = "labels",d = 30,
                    params=list(cydar=list(tol=tol_dataset[[data_id]], downsample=3)))
df <- .outcome_by_prob(cydar_res, 0.6) 

df

```



```{r}
sim.design <- model.matrix(design, data=design_df)[colnames(cd.dge),]
sim.dge <- estimateDisp(cd.dge, sim.design)
sim.fit <- glmQLFit(sim.dge, sim.design)
sim.res <- glmQLFTest(sim.fit, coef=2)

# control the spatial FDR
cydar.res <- sim.res$table
cydar.res$SpatialFDR <- spatialFDR(intensities(cd), sim.res$table$PValue)
boxplot(sim.dge$counts)
hist(cydar.res$PValue)

cydar.res$SpatialFDR < 0.1
```


```{r}
colData(sim_trajectory$SCE) <- DataFrame(sim_trajectory$meta)
plotPCA(sim_trajectory$SCE, colour_by="Condition")
plotReducedDim(sce, "pca.corrected",colour_by="synth_labels")
plotReducedDim(sce, "UMAP",colour_by="Condition1_prob")
```

Try downsampling one of the 2 conditions

```{r}
C1_cells <- colnames(sce)[sce$synth_labels=="Condition1"]
C2_cells <- colnames(sce)[sce$synth_labels=="Condition2"]
smp_C2 <- sample(C2_cells, size = length(C2_cells) * 1)

sce[,c(C1_cells, C2_cells)] 
sce
table(run_cydar(sce[,c(C1_cells, C2_cells)], tol = 3.0, d=30)$DAres$SpatialFDR < 0.1)
```
```{r}
benchmark_df <- read_csv("/nfs/team205/ed6/data/milo_benchmark/benchmark_embryo_pop_Erythroid2_enr0.9_seed43_batchEffect0.DAresults.daseq.csv") 

.outcome_by_prob(benchmark_df, 0.6)


.outcome_by_prob(benchmark_df, 0.6)
```


