---
title: "Batch effect - mouse chimeras"
output: html_notebook
---

```{r}
library(tidyverse)
library(MouseGastrulationData)
library(SingleCellExperiment)
library(scater)
library(scran)
library(bit64)

source('~/milo/notebooks/benchmark/benchmark_utils.R')
```

### Load dataset

Info from `MouseGastrulationAtlas` vignettes

> T (Brachyury) knockout chimeras involving sixteen samples, from eight embryo pools at two timepoints. The injected cells in the T chimeras have knockouts for the T gene. They also contain the td-Tomato construct. 

> They contain an extra feature for expression of the td-Tomato.
Cells derived from the injected cells (and thus are positive for td-Tomato) are marked in the colData field tomato.
Information for the proper pairing of samples from the same embryo pools can be found in the colData field pool.
Spliced count matrices are not provided.


```{r}
tchim_sce <- TChimeraData(type="processed")
wtchim_sce <- WTChimeraData(type="processed")

## Keep E8.5 samples
tchim_sce_e85 <- tchim_sce[,tchim_sce$stage == 'E8.5']
tchim_sce_e85 <- tchim_sce_e85[,!is.na(rowSums(reducedDim(tchim_sce_e85, "pca.corrected.E8.5")))]
wtchim_sce_e85 <- wtchim_sce[,wtchim_sce$stage == 'E8.5']
wtchim_sce_e85 <- wtchim_sce_e85[,!is.na(rowSums(reducedDim(wtchim_sce_e85, "pca.corrected.E8.5")))]

## Merge in one 
plotReducedDim(wtchim_sce_e85, dimred = 'pca.corrected.E8.5')
plotReducedDim(tchim_sce_e85, dimred = 'pca.corrected.E8.5')

all(rownames(tchim_sce_e85) %in% rownames(wtchim_sce_e85))
colData(tchim_sce_e85) <- colData(tchim_sce_e85)[colnames(colData(wtchim_sce_e85))]

chim_sce_e85 <- cbind(wtchim_sce_e85, tchim_sce_e85)
```

Attempt to reproduce batch correction from https://github.com/MarioniLab/TChimeras2020/blob/master/7_batch_correct/batch_correct.Rmd 

```{r}
#ensure counts has columns names for the cells
#match timepoints,samples to the count table
#timepoint_order, sample_order should contain each sample/timepoint ONCE, in correct order for correction
doBatchCorrect = function(counts, 
                          timepoints, 
                          samples, 
                          timepoint_order = c("E6.5", "E6.75", "E7.0", "mixed_gastrulation", "E7.25", "E7.5", "E7.75", "E8.0", "E8.25", "E8.5"), 
                          sample_order=NULL, 
                          npc = 50,
                          pc_override = NULL, 
                          BPPARAM = SerialParam()){
  require(scran)
  require(irlba)
  require(BiocParallel)
  
  #default: samples largest to smallest
  if(is.null(sample_order)){
    tab = table(samples)
    sample_order = names(tab)[order(tab), decreasing = TRUE]
  }
  
  #remove timepoints that are not present
  timepoint_order = timepoint_order[timepoint_order %in% timepoints]
  
  if(!is.null(pc_override)){
    pca = pc_override
  } else {
    pca = prcomp_irlba(t(counts), n = npc)$x
    rownames(pca) = colnames(counts)
  }
  
  if(length(unique(samples)) == 1){
    return(pca)
  }
  
  #create nested list
  pc_list = lapply(unique(timepoints), function(tp){
    sub_pc = pca[timepoints == tp, , drop = FALSE]
    sub_samp = samples[timepoints == tp]
    list = lapply(unique(sub_samp), function(samp){
      sub_pc[sub_samp == samp, , drop = FALSE]
    })
    names(list) = unique(sub_samp)
    return(list)
  })
  
  names(pc_list) = unique(timepoints)
  
  #arrange to match timepoint order
  pc_list = pc_list[order(match(names(pc_list), timepoint_order))]
  pc_list = lapply(pc_list, function(x){
    x[order(match(names(x), sample_order))]
  })
  
  #perform corrections within list elements (i.e. within stages)
  correct_list = lapply(pc_list, function(x){
    if(length(x) > 1){
      return(do.call(fastMNN, c(x, "pc.input" = TRUE, BPPARAM = BPPARAM))$corrected)
    } else {
      return(x[[1]])
    }
  })
  
  #perform correction over list
  if(length(correct_list)>1){
    correct = do.call(fastMNN, c(correct_list, "pc.input" = TRUE, BPPARAM = BPPARAM))$corrected
  } else {
    correct = correct_list[[1]]
  }
  
  correct = correct[match(colnames(counts), rownames(correct)),]
  
  return(correct)
  
}

## Get HVGs
chim_sce_e85 <- logNormCounts(chim_sce_e85)
dec <- modelGeneVar(chim_sce_e85)
hvgs_E8.5 <- getTopHVGs(dec, n=5000)

## Batch correct
meta <- as_tibble(colData(chim_sce_e85))
tab = table(meta$sample[meta$stage == "E8.5"])
correct2 = doBatchCorrect(counts = logcounts(chim_sce_e85), 
                          timepoints = meta$tomato[meta$stage == "E8.5"], 
                          samples = meta$sample[meta$stage == "E8.5"], 
                          timepoint_order = as.logical(c("TRUE", "FALSE")), 
                          sample_order = as.numeric(names(tab[order(tab, decreasing = TRUE)])))

```



```{r, fig.height=10, fig.width=10}
reducedDim(tchim_sce_e85, "PCA") <- reducedDim(tchim_sce_e85, "pca.corrected.E8.5")

tchim_sce_e85 <- runUMAP(tchim_sce_e85, dimred="PCA")

plot(
    x = reducedDim(tchim_sce_e85, "UMAP")[,1],
    y = reducedDim(tchim_sce_e85, "UMAP")[,2],
    col = EmbryoCelltypeColours[tchim_sce_e85$celltype.mapped],
    pch = 19,
    xaxt = "n", yaxt = "n",
    xlab = "UMAP1", ylab = "UMAP2"
)


```

```{r, fig.height=7, fig.width=8}
plotUMAP(tchim_sce_e85, colour_by="tomato", point_size=0.1) +
  facet_wrap(colour_by~.)
plotUMAP(tchim_sce_e85, colour_by="pool", point_size=0.1, text_by="celltype.mapped")
```

## Run Milo

W/o controlling for technical covariate

```{r}
## Make design matrix

sample_col = "sample"
design_df <- as.tibble(colData(tchim_sce_e85)[c(sample_col, "tomato", "pool")]) %>%
  distinct() %>%
  column_to_rownames(sample_col)

design <- formula("~ tomato")

## Build graph neighbourhoods
milo <- Milo(tchim_sce_e85)
milo <- buildGraph(milo, k=k, d=d, reduced.dim = reduced.dim)
milo <- makeNhoods(milo, prop = 0.1, k=15, d=30, reduced_dims = "pca.corrected.E8.5")

plotNhoodSizeHist(milo)
```
```{r}
## Test DA
milo <- countCells(milo, meta.data = data.frame(colData(milo)), sample=sample_col)
milo <- calcNhoodDistance(milo, d=30, reduced.dim = "pca.corrected.E8.5")
DA_results <- testNhoods(milo, design = design, design.df = design_df)

```
```{r, fig.width=10, fig.height=9}
milo <- buildNhoodGraph(milo)
plotNhoodGraphDA(milo, DA_results, alpha = 0.1)
```

