---
title: "Milo: liver cirrhosis analysis"
output: 
  html_notebook:
    code_folding: hide
---

## Introduction

In this notebook we demonstrate how to use Milo to detect abherrant cell states in diseased tissues, using a dataset of hepatic non-parenchymal cells isolated from 5 healthy and 5 cirrhotic human livers. [Ramachandran et al. 2019](https://www.nature.com/articles/s41586-019-1631-3#Sec1) (GEO accessiion: GSE136103).

```{r}
suppressPackageStartupMessages({
  library(tidyverse)
  library(irlba)
  library(DropletUtils)
  library(scater)
  library(scran)
  library(Seurat) ## just 4 loading the object
  library(miloR)
  library(SingleCellExperiment)
  library(patchwork)
  library(igraph)
  library(RColorBrewer)
  library(cowplot)
  })
```

## Load data

We downloaded the dataset and annotations stored in Seurat object from [here](https://datashare.is.ed.ac.uk/handle/10283/3433), as indicated by the authors.

```{r}
load("/nfs/team205/ed6/data/Ramachandran2019_liver/tissue.rdata")

## Convert to SingleCellExperiment
liver_sce <- SingleCellExperiment(assay = list(counts=tissue@raw.data, logcounts=tissue@data),
                                  colData = tissue@meta.data)

liver_sce
```

## Preprocessing

### Feature selection

```{r}
dec_liver <- modelGeneVar(liver_sce)

fit_liver <- metadata(dec_liver)
plot(fit_liver$mean, fit_liver$var, xlab="Mean of log-expression",
    ylab="Variance of log-expression")

hvgs <- getTopHVGs(dec_liver, n=3000)
```

### Dimensionality reduction with PCA

```{r, fig.height=14, fig.width=14}
liver_sce <- runPCA(liver_sce, subset_row=hvgs, ncomponents=11)

plotPCA(liver_sce, colour_by="condition", ncomponents=3)
```

```{r, fig.height=8, fig.width=8}
liver_sce <- runUMAP(liver_sce, dimred="PCA", ncomponents=2)

scater::plotUMAP(liver_sce, colour_by="condition", point_alpha=1,  point_size=0.5)
scater::plotUMAP(liver_sce, colour_by="dataset", point_alpha=0.3,  point_size=0.5)
scater::plotUMAP(liver_sce, colour_by="annotation_lineage", point_alpha=0.3,  point_size=0.5, text_by='annotation_lineage')
# scater::plotUMAP(liver_sce, colour_by='annotation_indepth', point_alpha=0.3,  point_size=0.5, text_by='annotation_indepth')
```

Notably, this dataset doesn't appear to display a batch effect

## DA analysis with Milo

We test for differential abundance between healthy and cirrhotic livers. We start by defining neighbourhoods with refined sampling on the KNN graph. We inspect the size of neighbourhoods.

```{r}
liver_milo <- Milo(liver_sce)

## Build KNN graph
liver_milo <- buildGraph(liver_milo, d = 11, k=30)

## Compute neighbourhoods with refined sampling
liver_milo <- makeNhoods(liver_milo, k=30, d=11, prop = 0.05, refined=TRUE)
plotNhoodSizeHist(liver_milo, bins=150)
```

Then we make a design matrix for the differential test, assigning samples to conditions.

```{r}
liver_meta <- as.tibble(colData(liver_milo)[,c("dataset","condition")])
liver_meta <- distinct(liver_meta) %>%
  mutate(condition=factor(condition, levels=c("Uninjured", "Cirrhotic"))) %>%
  column_to_rownames("dataset")
```

Now we can count cells in neighbourhoods and perform the DA test.

```{r}
liver_milo <- countCells(liver_milo, samples = "dataset", meta.data = data.frame(colData(liver_milo)[,c("dataset","condition")]) )

liver_milo <- calcNhoodDistance(liver_milo, d=11)
milo_res <- testNhoods(liver_milo, design = ~ condition, design.df = liver_meta, fdr.weighting = "k-distance")
```

## Exploration of DA results

We can start by looking at some basic stats

```{r}
pval_hist <- milo_res %>%
  ggplot(aes(PValue)) +
  geom_histogram(bins=50) +
  theme_bw(base_size=14)

volcano <-
  milo_res %>%
  ggplot(aes(logFC, -log10(SpatialFDR))) +
  geom_point(size=0.4, alpha=0.2) +
  geom_hline(yintercept = -log10(0.1)) +
  xlab("log-Fold Change") +
  theme_bw(base_size=14)

pval_hist + volcano
```

The distribution of P-values looks sensible and from the volcano plot we can see that we have identified some DA neighbourhoods at 10% FDR.
We can visualize DA neighbourhoods building an abstracted graph

```{r, fig.width=14, fig.height=10}
liver_milo <- buildNhoodGraph(liver_milo)
plotNhoodGraphDA(liver_milo, milo_res, alpha = 0.05)
```

```{r}
## Save milo object and results
saveRDS(liver_milo,"~/liver_milo_20201008.RDS")
write_csv(milo_res,"/nfs/team205/ed6/data/Ramachandran2019_liver/liver_results_20201008.csv")
```

```{r}
liver_milo <- readRDS("~/liver_milo_20201008.RDS")
milo_res <- read_csv("/nfs/team205/ed6/data/Ramachandran2019_liver/liver_results_20201008.csv")

## Load hvgs 
hvgs <- scan("~/liver_milo_hvgs.txt", "")
```



Making figures for the manuscript

```{r, fig.width=15, fig.height=10}
colourCount = length(unique(liver_milo$annotation_lineage))
getPalette = colorRampPalette(brewer.pal(9, "Set2"))

umap_df <- data.frame(reducedDim(liver_milo, "UMAP"))
colnames(umap_df) <- c("UMAP_1", "UMAP_2")

umap1 <- cbind(umap_df, annotation_lineage=liver_milo$annotation_lineage) %>%
  ggplot(aes(UMAP_1, UMAP_2, color=as.character(annotation_lineage))) +
  geom_point(size=0.1, alpha=0.5) +
  ggrepel::geom_text_repel(data = . %>%
              group_by(annotation_lineage) %>%
              summarise(UMAP_1=mean(UMAP_1), UMAP_2=mean(UMAP_2)),
            aes(label=annotation_lineage), color="black", size=6
            ) +
  scale_color_manual(values=getPalette(colourCount)) +
  guides(color="none") +
  xlab("UMAP1") + ylab("UMAP2") +
  coord_fixed() +
  theme_classic(base_size = 22) +
  theme(axis.text = element_blank(), axis.ticks = element_blank())

umap2 <-
  cbind(umap_df, condition=as.character(liver_milo$condition)) %>%
  ggplot(aes(UMAP_1, UMAP_2, color=condition)) +
  geom_point(size=0.1, alpha=0.5) +
  scale_color_brewer(palette="Set1", name='') +
  xlab("UMAP1") + ylab("UMAP2") +
  coord_fixed() +
  guides(color='none') +
  facet_wrap(condition~., ncol=1) +
  theme_nothing(font_size = 22) +
  theme(axis.text = element_blank(), axis.ticks = element_blank(), legend.position=c(0.9,0.9),
        strip.background = element_rect(color=NA), strip.text = element_text(size=22))

nh_graph_pl <- plotNhoodGraphDA(liver_milo, milo_res, alpha = 0.1) +
  theme(legend.text = element_text(size=22), legend.title = element_text(size=24)) +
  coord_fixed()

fig4_top <- (umap1 | umap2 | nh_graph_pl) +
  plot_layout(widths = c(3,1,3))

fig4_top +
  ggsave("~/milo_output/liver_embedding.pdf", width=15, height = 10)
```

Next, we can check the cell types where we observe most differences between healthy and cirrhotic cells, by taking the most frequent cell type in each neighbourhood.

```{r, fig.width=9, fig.height=10}
# Add annotation of most frequent cell type per nhood to milo results table
milo_res <- annotateNhoods(liver_milo, milo_res, "annotation_indepth")
anno_df <- data.frame(liver_milo@colData) %>%
  distinct(annotation_lineage, annotation_indepth)
milo_res <- left_join(milo_res, anno_df, by="annotation_indepth")
```

We first check that neighbourhoods are quite homogeneous

```{r}
frac_hist <- ggplot(milo_res, aes(annotation_indepth_fraction)) +
  geom_histogram(bins=30) +
  xlab("Fraction of cells in \nmost abundant cluster") +
  ylab("# neighbourhoods") +
  theme_bw(base_size=14)

frac_hist
```

I can recover all the clusters where DA was detected in the original paper (see all the barplots for each lineage) and more! All in a single analysis, and without knowing where the subclusters are. Let's bear in mind that positive logFC

```{r, fig.width=10, fig.height=10, warning=FALSE, message=FALSE}
group.by = "annotation_indepth"
paper_DA <- list(cirrhotic=c("MPs (4)","MPs (5)",
                             "Endothelia (6)", "Endothelia (7)",
                             "Mes (3)",
                             "Tcells (2)",
                             "Myofibroblasts"
                             ),
                 healthy=c("MPs (7)",
                           "Endothelia (1)",
                           "Tcells (1)", "Tcells (3)","Tcells (1)",
                           "ILCs (1)"
                           )
                 )

expDA_df <- bind_rows(
  data.frame(annotation_indepth = paper_DA[["cirrhotic"]], pred_DA="cirrhotic"),
  data.frame(annotation_indepth = paper_DA[["healthy"]], pred_DA="healthy")
  )

pl1 <- milo_res %>%
  left_join(expDA_df) %>%
  mutate(is_signif = ifelse(SpatialFDR < 0.1, 1, 0)) %>%
  mutate(logFC_color = ifelse(is_signif==1, logFC, NA)) %>%
  arrange(annotation_lineage) %>%
  mutate(Nhood=factor(Nhood, levels=unique(Nhood))) %>%
  ggplot(aes(annotation_indepth, logFC, color=logFC_color)) +
  scale_color_gradient2() +
  guides(color="none") +
  xlab(group.by) + ylab("Log Fold Change") +
  ggbeeswarm::geom_quasirandom(alpha=1) +
  coord_flip() +
  facet_grid(annotation_lineage~., scales="free", space="free") +
  theme_bw(base_size=22) +
  theme(strip.text.y =  element_text(angle=0),
        axis.title.y = element_blank(), axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        )

pl2 <- milo_res %>%
  left_join(expDA_df) %>%
  # dplyr::filter(!is.na(pred_DA)) %>%
  group_by(annotation_indepth) %>%
  summarise(pred_DA=dplyr::first(pred_DA), annotation_lineage=dplyr::first(annotation_lineage)) %>%
  mutate(end=ifelse(pred_DA=="healthy", 0, 1),
         start=ifelse(pred_DA=="healthy", 1, 0)) %>%
  ggplot(aes(annotation_indepth, start, xend = annotation_indepth, yend = end, color=pred_DA)) +
  geom_segment(size=1,arrow=arrow(length = unit(0.1, "npc"), type="closed")) +
  coord_flip() +
  xlab("annotation") +
  facet_grid(annotation_lineage~.,
    # annotation_lineage~"Ramachandran et al.\nDA predictions",
             scales="free", space="free") +
  # guides(color="none") +
  scale_color_brewer(palette="Set1", direction = -1,
                     labels=c("enriched in cirrhotic", "enriched in healthy"),
                     na.translate = F,
                     name="Ramachandran et al.\nDA predictions") +
  guides(color=guide_legend(ncol = 1)) +
  theme_bw(base_size=22) +
  ylim(-0.1,1.1) +
  theme(strip.text.y = element_blank(),strip.text.x = element_text(angle=90),
        plot.margin = unit(c(0,0,0,0), "cm"), panel.grid = element_blank(),
        axis.title.x = element_blank(), axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        legend.position = "bottom")

fig4_bleft <- (pl2 + pl1 +
  plot_layout(widths=c(1,10), guides = "collect") & theme(legend.position = 'top', legend.justification = 0))

fig4_bleft +
  ggsave("~/milo_output/liver_DAcomparison.pdf", width=8, height = 13)
```

## Close-up on Endothelial lineage

```{r}
endo_milo <- scater::runUMAP(liver_milo[,liver_milo$annotation_lineage=="Endothelia"],  dimred='PCA')
plotUMAP(endo_milo, colour_by = "annotation_indepth")
```

<!-- Filter out cells that show contamination from immune cells (expression of immune markers) -->

<!-- ```{r} -->
<!-- keep <- logcounts(endo_milo)["CD19",] == 0 | logcounts(endo_milo)["MS4A1",] == 0 -->
<!-- keep_dge <- logcounts(liver_milo)["CD19",] == 0 | logcounts(liver_milo)["MS4A1",] == 0 -->
<!-- endo_milo <- endo_milo[,keep] -->
<!-- endo_milo <- scater::runUMAP(endo_milo,  dimred='PCA') -->

<!-- plotUMAP(endo_milo, colour_by = "annotation_indepth") -->
<!-- ``` -->

```{r}
umap_df <- data.frame(reducedDim(endo_milo, "UMAP"))
colnames(umap_df) <- c("UMAP_1", "UMAP_2")

endo_umap <- cbind(umap_df, condition=endo_milo$condition) %>%
   ggplot(aes(UMAP_1, UMAP_2, color=condition)) +
  geom_point(size=0.3, alpha=0.5) +
  scale_color_brewer(palette="Set1", name='') +
  xlab("UMAP1") + ylab("UMAP2") +
  coord_fixed() +
  guides(color="none") +
  facet_wrap(condition~., ncol=1) +
  theme_nothing() +
  theme(axis.text = element_blank(), axis.ticks = element_blank(), legend.position=c(0.9,0.9),
        strip.background = element_rect(color=NA), strip.text = element_text(size=22))
```

```{r, fig.width=8, fig.height=4, message=FALSE, warning=FALSE}
liver_milo2 <- liver_milo
subset.nhoods <- str_detect(milo_res$annotation_indepth, "Endo")
reducedDim(liver_milo2, "UMAP")[colnames(endo_milo),] <- reducedDim(endo_milo, "UMAP") 

endo_gr <-
  plotNhoodGraphDA(
  liver_milo2, milo_res,
  subset.nhoods = subset.nhoods, 
  # ) =)[1:(length()-1)], 
  alpha = 0.1
  )  +
   theme(legend.text = element_text(size=22), legend.title = element_text(size=24))
  
fig4_bright1 <- ((endo_umap + endo_gr ) + 
  plot_layout(widths = c(1,2), 
                guides = "collect"
                )) 
fig4_bright1 +
  ggsave("~/milo_output/liver_endoGraph.pdf", width=9, height = 5)  
```

## Close-up on Cholangiocytes

```{r}
chol_milo <- scater::runUMAP(liver_milo[,liver_milo$annotation_lineage=="Cholangiocytes"],  dimred='PCA')
plotUMAP(chol_milo, colour_by = "annotation_indepth")

plotUMAP(chol_milo, colour_by = "percent.mito")
```

Filter out cells that show contamination from immune cells (expression of immune markers)

```{r}
keep <- logcounts(chol_milo)["CD19",] == 0 | logcounts(chol_milo)["MS4A1",] == 0
chol_milo <- chol_milo[,keep]
chol_milo <- scater::runUMAP(chol_milo,  dimred='PCA')

plotUMAP(chol_milo, colour_by = "annotation_indepth")
```

```{r, fig.width=10, fig.height=8}
umap_df <- data.frame(reducedDim(chol_milo, "UMAP"))
colnames(umap_df) <- c("UMAP_1", "UMAP_2")

chol_umap <- cbind(umap_df, condition=chol_milo$condition) %>%
   ggplot(aes(UMAP_1, UMAP_2, color=condition)) +
  geom_point(size=0.3, alpha=0.5) +
  scale_color_brewer(palette="Set1", name='') +
  xlab("UMAP1") + ylab("UMAP2") +
  coord_fixed() +
  guides(color="none") +
  facet_wrap(condition~., ncol=1) +
  theme_nothing() +
  theme(axis.text = element_blank(), axis.ticks = element_blank(), legend.position=c(0.9,0.9),
        strip.background = element_rect(color=NA), strip.text = element_text(size=22))

chol_umap
```

```{r, fig.width=8, fig.height=4, warning=FALSE, message=FALSE}
liver_milo2 <- liver_milo
subset.nhoods <- str_detect(milo_res$annotation_indepth, "Cholangio")
reducedDim(liver_milo2, "UMAP")[colnames(chol_milo),] <- reducedDim(chol_milo, "UMAP") 

chol_gr <-
  plotNhoodGraphDA(
  liver_milo2, milo_res,
  subset.nhoods = subset.nhoods, 
  # ) =)[1:(length()-1)], 
  alpha = 0.1
  )  +
   theme(legend.text = element_text(size=22), legend.title = element_text(size=24))
  
(chol_umap + chol_gr ) + 
  plot_layout(widths = c(1,2), 
                guides = "collect"
                )
# fig4_bright1 +
#   ggsave("~/milo_output/liver_endoGraph.pdf", width=9, height = 5)  

```

### DGE analysis


## Feature selection
```{r}
dec_liver <- modelGeneVar(liver_milo)
fit_liver <- metadata(dec_liver)
hvgs <- getTopHVGs(dec_liver, n=3000)
```

Add nhood expression to speed-up plotting of heatmaps
```{r}
liver_milo <- calcNhoodExpression(liver_milo, assay = "logcounts", subset.row = hvgs)
```


## Endothelia

<!-- ```{r} -->
<!-- subset.nhoods = milo_res$annotation_lineage=="Endothelia" -->
<!-- nhoodAdjacency(liver_milo) <- nhoodAdjacency(liver_milo)[names(nhoods(liver_milo)),names(nhoods(liver_milo))] -->
<!-- endo_markers <- findNhoodMarkers(liver_milo, milo_res, da.fdr = 0.1,aggregate.samples = TRUE, sample_col = "dataset", -->
<!--                  overlap = 1, lfc.threshold = 0.1, merge.discord = FALSE,  -->
<!--                  # subset.nhoods = milo_res$annotation_lineage=="Endothelia", -->
<!--                  subset.nhoods = as.character(unlist(nhoodIndex(liver_milo)[subset.nhoods])), -->
<!--                  subset.row = NULL, compute.new = FALSE, assay = "counts") -->
<!-- ``` -->
```{r}
x = liver_milo
da.res = milo_res
subset.row = hvgs
assay="counts"
aggregate.samples = TRUE
sample_col = "dataset"
gene.offset=TRUE

subset.nhoods = milo_res$annotation_lineage=="Endothelia"
overlap=1

nhood.adj <- nhoodAdjacency(liver_milo)[as.character(unlist(nhoodIndex(liver_milo)[subset.nhoods])),as.character(unlist(nhoodIndex(liver_milo)[subset.nhoods]))]

nhood.adj[which(milo_res[subset.nhoods, "logFC"] > 0),which(milo_res[subset.nhoods, "logFC"] < 0)] <- 0
nhood.adj[which(milo_res[subset.nhoods, "logFC"] < 0),which(milo_res[subset.nhoods, "logFC"] > 0)] <- 0

nhood.adj[nhood.adj < overlap] <- 0 
g <- graph_from_adjacency_matrix(nhood.adj)
nhs.da.gr <- components(g)$membership


nhood.gr <- unique(nhs.da.gr)
# perform DGE _within_ each group of cells using the input design matrix
message(paste0("Nhoods aggregated into ", length(nhood.gr), " groups"))

fake.meta <- data.frame("CellID"=colnames(x), "Nhood.Group"=rep(NA, ncol(x)))
rownames(fake.meta) <- fake.meta$CellID

# do we want to allow cells to be members of multiple groups? This will create
# chaos for the LM as there will be a dependency structure comparing 2 different
# groups that contain overlapping cells.
# this approach means that the latter group takes precedent.
# maybe exclude the cells that fall into separate groups?
for(i in seq_along(nhood.gr)){
    nhood.x <- nhs.da.gr == nhood.gr[i]
    # get the nhoods
    nhs <- nhoods(x)
    if(!is.null(subset.nhoods)){
        nhs <- nhs[subset.nhoods]
    }

    if(!any(is.na(fake.meta[unlist(nhs[nhood.x]),]$Nhood.Group))){
        fake.meta[unlist(nhs[nhood.x]),]$Nhood.Group[!is.na(fake.meta[unlist(nhs[nhood.x]),]$Nhood.Group)] <- NA
        } else{
            fake.meta[unlist(nhs[nhood.x]),]$Nhood.Group <- nhood.gr[i]
        }
}

# only compare against the other DA neighbourhoods
x <- x[, !is.na(fake.meta$Nhood.Group)]
fake.meta <- fake.meta[!is.na(fake.meta$Nhood.Group), ]

if(!is.null(subset.row)){
    x <- x[subset.row, , drop=FALSE]
}

exprs <- assay(x, assay)

marker.list <- list()
i.contrast <- c("TestTest - TestRef") # always use contrasts for this

    # if there is only 1 group, then need to make sure that all neighbourhoods
    # are not in this group - otherwise can't do any DGE testing
    if(length(nhood.gr) == 1){
        if(sum(fake.meta$Nhood.Group == nhood.gr[1]) == nrow(fake.meta)){
            warning("All graph neighbourhoods are in the same group - cannot perform DGE testing. Returning NULL")
            return(NULL)
        }
    }
    
    ## Aggregate expression by sample
    # To avoid treating cells as independent replicates
    if (isTRUE(aggregate.samples)) {
        fake.meta[,"sample_id"] <- colData(x)[[sample_col]]
        fake.meta[,'sample_group'] <- paste(fake.meta[,"sample_id"], fake.meta[,"Nhood.Group"], sep="_")
        
        sample_gr_mat <- matrix(0, nrow=nrow(fake.meta), ncol=length(unique(fake.meta$sample_group)))
        colnames(sample_gr_mat) <- unique(fake.meta$sample_group)
        rownames(sample_gr_mat) <- rownames(fake.meta)
        
        for (s in colnames(sample_gr_mat)) {
            sample_gr_mat[which(fake.meta$sample_group == s),s] <- 1  
        }
        
        ## Summarise expression by sample
        exprs_smp <- matrix(0, nrow=nrow(exprs), ncol=ncol(sample_gr_mat))
        if (assay=='counts') {
            summFunc <- rowSums
        } else {
            summFunc <- rowMeans
        }
        
        for (i in 1:ncol(sample_gr_mat)){
            if (sum(sample_gr_mat[,i]) > 1) {
                exprs_smp[,i] <- summFunc(exprs[,which(sample_gr_mat[,i] > 0)])  
            } else {
                exprs_smp[,i] <- exprs[,which(sample_gr_mat[,i] > 0)]
            }
        }
        rownames(exprs_smp) <- rownames(exprs)
        colnames(exprs_smp) <- colnames(sample_gr_mat)
        
        smp_meta <- distinct(fake.meta, sample_group, Nhood.Group)
        rownames(smp_meta) <- smp_meta[,"sample_group"]
        
        fake.meta <- smp_meta
        exprs <- exprs_smp
    }
    
    for(i in seq_along(nhood.gr)){
        i.meta <- fake.meta
        i.meta$Test <- "Ref"
        i.meta$Test[fake.meta$Nhood.Group == nhood.gr[i]] <- "Test"

        if(ncol(exprs) > 1 & nrow(i.meta) > 1){
            i.design <- as.formula(" ~ 0 + Test")
            i.model <- model.matrix(i.design, data=i.meta)
            rownames(i.model) <- rownames(i.meta)
        }

        sink(file="/dev/null")
        gc()
        sink(file=NULL)

        if(assay == "logcounts"){
            i.res <- .perform_lognormal_dge(exprs, i.model, model.contrasts=i.contrast,
                                            gene.offset=gene.offset)
        } else if(assay == "counts"){
            i.res <- .perform_counts_dge(exprs, i.model, model.contrasts=i.contrast,
                                         gene.offset=gene.offset)
            colnames(i.res)[ncol(i.res)] <- "adj.P.Val"
        } else{
            warning("Assay type is not counts or logcounts - assuming (log)-normal distribution. Use these results at your peril")
            i.res <- .perform_lognormal_dge(exprs, i.model,
                                            model.contrasts=i.contrast,
                                            gene.offset=gene.offset)
        }

        i.res$adj.P.Val[is.na(i.res$adj.P.Val)] <- 1
        i.res$logFC[is.infinite(i.res$logFC)] <- 0

        i.res <- i.res[, c("logFC", "adj.P.Val")]
        colnames(i.res) <- paste(colnames(i.res), nhood.gr[i], sep="_")
        marker.list[[paste0(nhood.gr[i])]] <- i.res
    }

    marker.df <- do.call(cbind.data.frame, marker.list)
    colnames(marker.df) <- gsub(colnames(marker.df), pattern="^[0-9]+\\.", replacement="")
    marker.df$GeneID <- rownames(i.res)
```

```{r}
marker.df[marker.df$adj.P.Val_2 < 0.1,]
```

#### Visualize as volcano 

```{r, fig.height=6, fig.width=10, message=FALSE, warning=FALSE}
marker.df %>%
  mutate(label=ifelse((adj.P.Val_2) < 0.01, GeneID, NA)) %>%
  ggplot(aes(logFC_2, -log10(adj.P.Val_2), 
             # color=highlight
             )) + 
  geom_point() +
  ggrepel::geom_text_repel(aes(label=label)) +
  xlab("logFC") + ylab("- log10(Adj. P value)") +
  theme_bw(base_size = 22)

marker.df %>%
  mutate(label=ifelse((adj.P.Val_1) < 0.01, GeneID, NA)) %>%
  ggplot(aes(logFC_1, -log10(adj.P.Val_1), 
             # color=highlight
             )) + 
  geom_point() +
  ggrepel::geom_text_repel(aes(label=label)) +
  xlab("logFC") + ylab("- log10(Adj. P value)") +
  theme_bw(base_size = 22)

  
```

#### Visualize as heatmap 
(gene expression values are scaled between 0 and 1 for each gene)

```{r, fig.height=9, fig.width=12, message=FALSE, warning=FALSE}
marker_genes <- marker.df %>%
  dplyr::filter(adj.P.Val_2 < 0.1) %>%
  pull(GeneID)

highlight_genes <- c("PLVAP", "VWA1", "ACKR1", "IL32",
                     "CLEC4G", "CLEC4M", "FCN2", "FCN3",
                     "LEF1")

fig4_bbright <- plotNhoodExpressionDA(liver_milo, milo_res, marker_genes, cluster_features = TRUE, 
                      alpha = 0.1,
                      scale_to_1 = TRUE,
                      subset.nhoods = milo_res$annotation_lineage=="Endothelia",
                      highlight_features = highlight_genes, show_rownames = FALSE
                      ) +
  ylab("DE genes") +
   theme(legend.text = element_text(size=22), legend.title = element_text(size=24)) +
  plot_layout(heights = c(1,10)) & theme(legend.margin = margin(0,0,0,60), legend.background = element_blank())

fig4_bbright
```

### GO term analysis

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# BiocManager::install('clusterProfiler')
# BiocManager::install('msigdbr')
library(clusterProfiler)
library(msigdbr)

m_df <- msigdbr(species = "Homo sapiens")
m_t2g <- msigdbr(species = "Homo sapiens", category = "C5", subcategory = "BP")  %>% 
  dplyr::select(gs_name, gene_symbol)

marker_genes_up <- marker.df %>%
  dplyr::filter(adj.P.Val_2 < 0.1 & logFC_1 < 0) %>%
  pull(GeneID)

marker_genes_down <- marker.df %>%
  dplyr::filter(adj.P.Val_1 < 0.1 & logFC_1 < 0) %>%
  pull(GeneID)

em_up <- enricher(marker_genes_up, TERM2GENE=m_t2g, pAdjustMethod = "fdr", 
                  universe = rownames(liver_milo)
                  )
em_down <- enricher(marker_genes_down, TERM2GENE=m_t2g, pAdjustMethod = "fdr", 
                    universe = rownames(liver_milo)
                    )

em_res_up <- em_up@result[em_up@result$qvalue < 0.1,] %>%
  dplyr::select(- c(Description))
em_res_down <- em_down@result[em_down@result$qvalue < 0.1,] %>%
  dplyr::select(- c(Description))
```

```{r, fig.height=8, fig.width=15, warning=FALSE, message=FALSE}
go_endo_up <- em_res_up %>%
  top_n(20, -log10(qvalue)) %>%
  mutate(Term=factor(ID, levels=rev(unique(ID)))) %>%
  ggplot(aes(Term, -log10(qvalue))) +
  geom_point() +
  coord_flip() +
  xlab("GO Biological Function") + ylab("-log10(Adj. p-value)") +
  theme_bw(base_size=18) +
  ggtitle("Cirrhotic endothelia")

go_endo_down <- em_res_down %>%
  top_n(20, -log10(qvalue)) %>%
  mutate(ID=ifelse(ID=='GO_ANTIGEN_PROCESSING_AND_PRESENTATION_OF_PEPTIDE_OR_POLYSACCHARIDE_ANTIGEN_VIA_MHC_CLASS_II', "GO_ANTIGEN_PRESENTATION_VIA_MHC_CLASS_II", ID)) %>%
  mutate(Term=factor(ID, levels=rev(unique(ID)))) %>%
  ggplot(aes(Term, -log10(qvalue))) +
  geom_point() +
  coord_flip() +
  xlab("GO Biological Function") + ylab("-log10(Adj. p-value)") +
  theme_bw(base_size=18) +
  ggtitle("Uninjured endothelia")

go_endo_up
go_endo_down
```


```{r}
em_res_up
em_res_down
```

### Cholangiocytes
```{r}
x = liver_milo
da.res = milo_res
subset.row = hvgs
assay="counts"
aggregate.samples = TRUE
sample_col = "dataset"
gene.offset=TRUE

subset.nhoods = milo_res$annotation_lineage=="Cholangiocytes"
overlap=1

nhood.adj <- nhoodAdjacency(liver_milo)[as.character(unlist(nhoodIndex(liver_milo)[subset.nhoods])),as.character(unlist(nhoodIndex(liver_milo)[subset.nhoods]))]

nhood.adj[which(milo_res[subset.nhoods, "logFC"] > 0),which(milo_res[subset.nhoods, "logFC"] < 0)] <- 0
nhood.adj[which(milo_res[subset.nhoods, "logFC"] < 0),which(milo_res[subset.nhoods, "logFC"] > 0)] <- 0

nhood.adj[nhood.adj < overlap] <- 0 
g <- graph_from_adjacency_matrix(nhood.adj)
nhs.da.gr <- components(g)$membership


nhood.gr <- unique(nhs.da.gr)
# perform DGE _within_ each group of cells using the input design matrix
message(paste0("Nhoods aggregated into ", length(nhood.gr), " groups"))

fake.meta <- data.frame("CellID"=colnames(x), "Nhood.Group"=rep(NA, ncol(x)))
rownames(fake.meta) <- fake.meta$CellID

# do we want to allow cells to be members of multiple groups? This will create
# chaos for the LM as there will be a dependency structure comparing 2 different
# groups that contain overlapping cells.
# this approach means that the latter group takes precedent.
# maybe exclude the cells that fall into separate groups?
for(i in seq_along(nhood.gr)){
    nhood.x <- nhs.da.gr == nhood.gr[i]
    # get the nhoods
    nhs <- nhoods(x)
    if(!is.null(subset.nhoods)){
        nhs <- nhs[subset.nhoods]
    }

    if(!any(is.na(fake.meta[unlist(nhs[nhood.x]),]$Nhood.Group))){
        fake.meta[unlist(nhs[nhood.x]),]$Nhood.Group[!is.na(fake.meta[unlist(nhs[nhood.x]),]$Nhood.Group)] <- NA
        } else{
            fake.meta[unlist(nhs[nhood.x]),]$Nhood.Group <- nhood.gr[i]
        }
}

# only compare against the other DA neighbourhoods
x <- x[, !is.na(fake.meta$Nhood.Group)]
fake.meta <- fake.meta[!is.na(fake.meta$Nhood.Group), ]

if(!is.null(subset.row)){
    x <- x[subset.row, , drop=FALSE]
}

exprs <- assay(x, assay)

marker.list <- list()
i.contrast <- c("TestTest - TestRef") # always use contrasts for this

    # if there is only 1 group, then need to make sure that all neighbourhoods
    # are not in this group - otherwise can't do any DGE testing
    if(length(nhood.gr) == 1){
        if(sum(fake.meta$Nhood.Group == nhood.gr[1]) == nrow(fake.meta)){
            warning("All graph neighbourhoods are in the same group - cannot perform DGE testing. Returning NULL")
            return(NULL)
        }
    }
    
    ## Aggregate expression by sample
    # To avoid treating cells as independent replicates
    if (isTRUE(aggregate.samples)) {
        fake.meta[,"sample_id"] <- colData(x)[[sample_col]]
        fake.meta[,'sample_group'] <- paste(fake.meta[,"sample_id"], fake.meta[,"Nhood.Group"], sep="_")
        
        sample_gr_mat <- matrix(0, nrow=nrow(fake.meta), ncol=length(unique(fake.meta$sample_group)))
        colnames(sample_gr_mat) <- unique(fake.meta$sample_group)
        rownames(sample_gr_mat) <- rownames(fake.meta)
        
        for (s in colnames(sample_gr_mat)) {
            sample_gr_mat[which(fake.meta$sample_group == s),s] <- 1  
        }
        
        ## Summarise expression by sample
        exprs_smp <- matrix(0, nrow=nrow(exprs), ncol=ncol(sample_gr_mat))
        if (assay=='counts') {
            summFunc <- rowSums
        } else {
            summFunc <- rowMeans
        }
        
        for (i in 1:ncol(sample_gr_mat)){
            if (sum(sample_gr_mat[,i]) > 1) {
                exprs_smp[,i] <- summFunc(exprs[,which(sample_gr_mat[,i] > 0)])  
            } else {
                exprs_smp[,i] <- exprs[,which(sample_gr_mat[,i] > 0)]
            }
        }
        rownames(exprs_smp) <- rownames(exprs)
        colnames(exprs_smp) <- colnames(sample_gr_mat)
        
        smp_meta <- distinct(fake.meta, sample_group, Nhood.Group)
        rownames(smp_meta) <- smp_meta[,"sample_group"]
        
        fake.meta <- smp_meta
        exprs <- exprs_smp
    }
    
    
    for(i in seq_along(nhood.gr)){
        i.meta <- fake.meta
        i.meta$Test <- "Ref"
        i.meta$Test[fake.meta$Nhood.Group == nhood.gr[i]] <- "Test"

        if(ncol(exprs) > 1 & nrow(i.meta) > 1){
            i.design <- as.formula(" ~ 0 + Test")
            i.model <- model.matrix(i.design, data=i.meta)
            rownames(i.model) <- rownames(i.meta)
        }

        sink(file="/dev/null")
        gc()
        sink(file=NULL)

        if(assay == "logcounts"){
            i.res <- .perform_lognormal_dge(exprs, i.model, model.contrasts=i.contrast,
                                            gene.offset=gene.offset)
        } else if(assay == "counts"){
            i.res <- .perform_counts_dge(exprs, i.model, model.contrasts=i.contrast,
                                         gene.offset=gene.offset)
            colnames(i.res)[ncol(i.res)] <- "adj.P.Val"
        } else{
            warning("Assay type is not counts or logcounts - assuming (log)-normal distribution. Use these results at your peril")
            i.res <- .perform_lognormal_dge(exprs, i.model,
                                            model.contrasts=i.contrast,
                                            gene.offset=gene.offset)
        }

        i.res$adj.P.Val[is.na(i.res$adj.P.Val)] <- 1
        i.res$logFC[is.infinite(i.res$logFC)] <- 0

        i.res <- i.res[, c("logFC", "adj.P.Val")]
        colnames(i.res) <- paste(colnames(i.res), nhood.gr[i], sep="_")
        marker.list[[paste0(nhood.gr[i])]] <- i.res
    }

    marker.df.chol <- do.call(cbind.data.frame, marker.list)
    colnames(marker.df.chol) <- gsub(colnames(marker.df.chol), pattern="^[0-9]+\\.", replacement="")
    marker.df.chol$GeneID <- rownames(i.res)
```
```{r}
table(nhs.da.gr)
```

```{r}
marker.df.chol[marker.df.chol$adj.P.Val_1 < 0.1,]
```

#### Visualize as volcano 

```{r, fig.height=6, fig.width=10, message=FALSE, warning=FALSE}
volcano_chol <- marker.df.chol %>%
  mutate(label=ifelse((adj.P.Val_1) < 0.1, GeneID, NA)) %>%
  ggplot(aes(logFC_1, -log10(adj.P.Val_1), 
             # color=highlight
             )) + 
  geom_point() +
  ggrepel::geom_text_repel(aes(label=label), segment.alpha = 0.3) +
  xlab("logFC") + ylab("- log10(Adj. P value)") +
  theme_bw(base_size = 22)

volcano_chol  
  
```

#### Visualize as heatmap 
(gene expression values are scaled between 0 and 1 for each gene)

```{r, fig.height=10, fig.width=9, message=FALSE, warning=FALSE}
marker_genes_chol <- marker.df.chol %>%
  dplyr::filter(adj.P.Val_1 < 0.1) %>%
  pull(GeneID)

highlight_genes <- c("PLVAP", "SOX14", "VWA1", "ACKR1", "IL32",
                     "CLEC4G", "CLEC4M", "STAB2", "MRC1", "LYVE1",
                     "CD14", "SOX17", "WNT2", "RSPO3", "AIF1L",
                     "PROX1", "PDPN","CPE", "CD320", "BMPER")

plotNhoodExpressionDA(liver_milo, milo_res, marker_genes_chol, cluster_features = TRUE, 
                      alpha = 0.1,
                      scale_to_1 = TRUE,
                      subset.nhoods = milo_res$annotation_lineage=="Cholangiocytes",
                      highlight_features = highlight_genes, show_rownames = TRUE
                      ) +
  ylab("DE genes") +
   theme(legend.text = element_text(size=22), legend.title = element_text(size=24)) +
  plot_layout(heights = c(1,10))


```


### GO term analysis

```{r, echo=FALSE, warning=FALSE, message=FALSE}
marker_genes_chol <- marker.df.chol %>%
  dplyr::filter(adj.P.Val_2 < 0.1 & logFC_2 < 0) %>%
  pull(GeneID)

em_up_chol <- enricher(marker_genes_chol, TERM2GENE=m_t2g, pAdjustMethod = "fdr", 
                  universe = rownames(liver_milo)
                  )

em_res_up_chol <- em_up_chol@result[em_up_chol@result$qvalue < 0.1,] %>%
  dplyr::select(- c(Description))
```

```{r, fig.height=8, fig.width=15, warning=FALSE, message=FALSE}
go_chol_up <- em_res_up_chol %>%
  top_n(20, -log10(qvalue)) %>%
  mutate(Term=factor(ID, levels=rev(unique(ID)))) %>%
  ggplot(aes(Term, -log10(qvalue))) +
  geom_point() +
  coord_flip() +
  xlab("GO Biological Function") + ylab("-log10(Adj. p-value)") +
  theme_bw(base_size=18) +
  ggtitle("Cirrhotic cholangiocytes")

go_chol_up
```

```{r}
em_res_up_chol
```


---

Assemble figure
```{r, fig.height=25, fig.width=19}
fig4_bottom <- ((fig4_bleft + plot_layout()) |
      ((fig4_bright1 + plot_layout(tag_level = 'keep')) / (fig4_bbright + plot_layout())) +
      plot_layout(heights = c(1,1.6))
   ) +
  plot_layout(widths=c(1,1.2))

(fig4_top / fig4_bottom) +
  plot_layout(heights=c(1,1.8))  +
  ggsave("~/milo_output/fig4.pdf", height = 26, width = 22, useDingbats=FALSE)
  # ggsave("~/milo/ms/figures/figs/figure5.pdf", height = 26, width = 22, useDingbats=FALSE)
```

Assemble supplementary figure

```{r, fig.width=25, fig.height=7}
p1 <- plot_grid( go_endo_up+ theme(plot.title = element_text(hjust = 1),
                                   axis.title.x = element_text(hjust = 1)), 
                 go_endo_down+ theme(plot.title = element_text(hjust = 1),
                                     axis.title.x = element_text(hjust = 1)), 
                 label_size = 18,
                 ncol=1,
                 rel_heights = c(2,2),
                labels = c("A", "B","C"))

p1

chol_emb <- (chol_umap + chol_gr ) + 
  plot_layout(widths = c(1,2), 
                guides = "collect"
                )

p3 <- plot_grid(plot_grid(chol_umap, chol_gr, nrow=1,rel_widths = c(1,2),
                          label_size = 18,
                labels = c("C","D")),
                volcano_chol,
                go_chol_up + theme(plot.title = element_text(hjust = 1),
                                   axis.title.x = element_text(hjust = 1)), 
                ncol=1,
                rel_heights = c(1.5,1,1.5),
                 label_size = 18,
                labels=c("","E",'F'))

plot_grid(p1, p3, ncol=2, rel_widths = c(1,1.2)) +
# cowplot::plot_grid(p1,p2,go_chol_up,  rel_heights = c(1,1.5), ncol=1, labels = c("", "F")) +
  ggsave("~/milo/ms/supplement/suppl_figs/suppl_fig6.pdf", height = 16, width=20)
  ggsave("~/milo/ms/supplement/suppl_figs/suppl_fig6.png", height = 16, width=20)
  ggsave("~/milo_output/liver_supplementary.png", height = 7, width=8)

```

<!-- Let's check the genes identified as markers for the disease subtypes. Are they significantly differentially expressed between DA neighbourhoods? Yes they are! -->

<!-- ```{r} -->
<!-- disease_endo_markers <- c("ACKR1", 'CD34',"VWA1") -->

<!-- data.frame(markers$negLogFC_2) %>% -->
<!--   filter(FDR < 0.05) %>% -->
<!--   .[disease_endo_markers,] -->

<!-- data.frame(markers$negLogFC_1) %>% -->
<!--   filter(FDR < 0.05) %>% -->
<!--   .[disease_endo_markers,] -->
<!-- ``` -->

<!-- Visualize some of the markers that differentiate DA neighbourhoods, by plotting the percent of cells expressing each gene in a nhood. -->

<!-- ```{r} -->
<!-- ## Define plotting functions -->
<!-- .calculate_nhood_perc_expression <- function(milo, nhoods, gene){ -->
<!--   gene_cnts <- counts(milo)[gene,] -->
<!--   perc_expr <- sapply(nhoods(milo)[nhoods], function(x) sum(gene_cnts[x]>0)/length(x)) -->
<!--   perc_expr <- setNames(perc_expr, nhoods) -->
<!--   return(perc_expr) -->
<!--   } -->

<!-- .plot_nhood_expression <- function(milo, nhoods, features){ -->
<!--   perc_expr_mat <- sapply(features,  -->
<!--                           function(x) .calculate_nhood_perc_expression(milo, nhoods, x)) -->

<!--   pl_df <- data.frame(perc_expr_mat) %>% -->
<!--     rownames_to_column("Nhood") %>% -->
<!--     mutate(Nhood=as.double(Nhood)) %>% -->
<!--     left_join(milo_res) %>% -->
<!--     mutate(logFC_rank=rank(logFC))  -->

<!--   pl_top <- pl_df %>% -->
<!--       mutate(is_signif = ifelse(SpatialFDR < 0.1, "SpatialFDR < 0.1", NA)) %>% -->
<!--       ggplot(aes(logFC_rank, logFC)) + -->
<!--       geom_hline(yintercept = 0, linetype=2) + -->
<!--       geom_point(size=0.2) + -->
<!--       geom_point(data=.%>% filter(!is.na(is_signif)), aes(color=is_signif), size=0.5) + -->
<!--       theme_bw() + -->
<!--       scale_color_manual(values="red", name="") + -->
<!--       theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(), axis.title.x = element_blank()) -->

<!--   pl_bottom <- pl_df %>% -->
<!--     pivot_longer(cols=features, names_to='feature', values_to="perc_expressed") %>% -->
<!--     mutate(feature=factor(feature, levels=features)) %>% -->
<!--     ggplot(aes(logFC_rank, feature, fill=perc_expressed)) +  -->
<!--     geom_tile() + -->
<!--     scale_fill_viridis_c(option="magma") + -->
<!--     ggbio::theme_clear() -->

<!--   (pl_top / pl_bottom) + plot_layout(heights = c(1,2)) -->
<!-- } -->
<!-- ``` -->


<!-- ```{r, fig.width=10, fig.height=12} -->
<!-- endo_nhoods <- endo_res %>%  pull(Nhood) -->

<!-- ## Select genes and sort by AUC -->
<!-- feats_neg2 <- -->
<!--   data.frame(markers$negLogFC_2) %>%  -->
<!--   top_n(50, - log10(FDR)) %>% -->
<!--   arrange(AUC.posLogFC_1) %>% -->
<!--   rownames() -->

<!-- .plot_nhood_expression(liver_milo, endo_nhoods, features=feats) -->
<!-- ``` -->

<!-- As described in the paper, we have that genes associated with extracellular matrix organization (e.g. VIM, ) are over expressed  -->


<!-- ```{r, fig.width=10, fig.height=7} -->
<!-- ## Select genes and sort by AUC -->
<!-- feats_neg1 <- -->
<!--   data.frame(markers$negLogFC_1) %>%  -->
<!--   top_n(50, - log10(FDR)) %>% -->
<!--   arrange(AUC.posLogFC_1) %>% -->
<!--   rownames() -->

<!-- .plot_nhood_expression(liver_milo, endo_nhoods, features=feats) -->

<!-- ``` -->

<!-- ```{r, fig.width=10, fig.height=7} -->
<!-- feats_neg1vsNeg2 <- -->
<!--   data.frame(markers$negLogFC_1) %>%  -->
<!--   top_n(50, - log10(FDR)) %>% -->
<!--   arrange(AUC.negLogFC_2) %>% -->
<!--   rownames() -->

<!-- .plot_nhood_expression(liver_milo, endo_nhoods, features=feats_neg1vsNeg2) -->
<!-- ``` -->


<!-- Look just at Endothelia (5) where you have both positive and negative fold-changes -->

<!-- ```{r} -->
<!-- endo5_nhoods <- milo_res %>%  -->
<!--   filter(annotation_indepth=="Endothelia (5)" & annotation_indepth_fraction > 0.7) %>% -->
<!--   pull(Nhood) -->

<!-- perc_expr_mat <- sapply(features,  -->
<!--                         function(x) .calculate_nhood_perc_expression(liver_milo, endo5_nhoods, x)) -->

<!-- pl_df <- data.frame(perc_expr_mat) %>% -->
<!--   rownames_to_column("Nhood") %>% -->
<!--   mutate(Nhood=as.double(Nhood)) %>% -->
<!--   left_join(milo_res) %>% -->
<!--   mutate(logFC_rank=rank(logFC))  -->

<!-- pl_top <- pl_df %>% -->
<!--     mutate(is_signif = ifelse(SpatialFDR < 0.1, "SpatialFDR < 0.1", NA)) %>% -->
<!--     ggplot(aes(logFC_rank, logFC)) + -->
<!--     geom_hline(yintercept = 0, linetype=2) + -->
<!--     geom_point(size=0.2) + -->
<!--     geom_point(data=.%>% filter(!is.na(is_signif)), aes(color=is_signif), size=0.5) + -->
<!--     theme_bw() + -->
<!--     scale_color_manual(values="red", name="") + -->
<!--     theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(), axis.title.x = element_blank()) -->

<!-- pl_bottom <- pl_df %>% -->
<!--   pivot_longer(cols=features, names_to='feature', values_to="perc_expressed") %>% -->
<!--   ggplot(aes(logFC_rank, feature, fill=perc_expressed)) +  -->
<!--   geom_tile() + -->
<!--   scale_fill_viridis_c(option="magma") + -->
<!--   theme_clear() -->

<!-- (pl_top / pl_bottom) + plot_layout(heights = c(1,2)) -->
<!-- ``` -->

<!-- How to find association de novo in Endo 5? -->

<!-- ```{r, fig.height=8, fig.width=8} -->
<!-- ## Find most highly variable genes in this cluster -->
<!-- endo5_milo <- liver_milo[,which(colData(liver_milo)[["annotation_indepth"]]=="Endothelia (5)")] -->
<!-- dec_endo5 <- modelGeneVar(endo5_milo) -->
<!-- endo5_hvgs <- getTopHVGs(dec_endo5, n=1000) -->

<!-- perc_expr_mat <- sapply(endo5_hvgs, function(x) .calculate_nhood_perc_expression(liver_milo, endo5_nhoods, x)) -->

<!-- milo_res_endo5 <- milo_res[which(milo_res$Nhood %in% endo5_nhoods),] -->

<!-- fc_cor <- apply(perc_expr_mat, 2, function(x) Hmisc::rcorr(x, milo_res_endo5$logFC)$r[1,2]) -->
<!-- fc_cor_pval <- apply(perc_expr_mat, 2, function(x) Hmisc::rcorr(x, milo_res_endo5$logFC)$P[1,2]) -->

<!-- cor_feats <- names(which(abs(fc_cor) > 0.6 & abs(fc_cor_pval) < 0.05)) -->
<!-- cor_feats_ordered <- cor_feats[order(fc_cor[cor_feats])] -->

<!-- pl_df <- data.frame(perc_expr_mat[,cor_feats]) %>% -->
<!--   rownames_to_column("Nhood") %>% -->
<!--   mutate(Nhood=as.double(Nhood)) %>% -->
<!--   left_join(milo_res) %>% -->
<!--   mutate(logFC_rank=rank(logFC))  -->

<!-- pl_top <- pl_df %>% -->
<!--     mutate(is_signif = ifelse(SpatialFDR < 0.1, "SpatialFDR < 0.1", NA)) %>% -->
<!--     ggplot(aes(logFC_rank, logFC)) + -->
<!--     geom_hline(yintercept = 0, linetype=2) + -->
<!--     geom_point(size=0.2) + -->
<!--     geom_point(data=.%>% filter(!is.na(is_signif)), aes(color=is_signif), size=0.5) + -->
<!--     theme_bw() + -->
<!--     scale_color_manual(values="red", name="") + -->
<!--     theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(), axis.title.x = element_blank()) -->

<!-- pl_bottom <- pl_df %>% -->
<!--   pivot_longer(cols=str_replace(cor_feats_ordered, "-", "."), names_to='feature', values_to="perc_expressed") %>% -->
<!--   mutate(feature=factor(feature, levels=str_replace(cor_feats_ordered, "-", "."))) %>% -->
<!--   ggplot(aes(logFC_rank, feature, fill=perc_expressed)) +  -->
<!--   geom_tile() + -->
<!--   scale_fill_viridis_c(option="magma") + -->
<!--   theme_clear() -->

<!-- (pl_top / pl_bottom) + plot_layout(heights = c(1,2)) -->
<!-- ``` -->



<!-- ```{r} -->
<!-- ## Run common PCA -->
<!-- merged_cnts <- cbind(nhoodExpression(endo_milo), logcounts(endo_milo)[hvgs,]) -->
<!-- merged_cnts_scaled <- t(scale(t(merged_cnts))) -->
<!-- merged_pca <- BiocSingular::runPCA(t(merged_cnts_scaled), rank=30, center=FALSE) -->
<!-- pca_mat <- rbind(merged_pca$x[(ncol(endo_milo)+1):(ncol(endo_milo)+(length(nhoods(endo_milo)))),], merged_pca$x[colnames(endo_milo),]) -->
<!-- ## Add to slot nhoodsReducedDim -->
<!-- nhoodReducedDim(endo_milo, "PCA") <- pca_mat -->

<!-- ## Run UMAP on joint PCA -->
<!-- umap_out <- uwot::umap(nhoodReducedDim(endo_milo, "PCA"), n_neighbors = 20, n_components = 2, scale=FALSE) -->
<!-- colnames(umap_out) <- c("UMAP_1", "UMAP_2") -->
<!-- nhoodReducedDim(endo_milo, "UMAP") <- umap_out -->

<!-- ``` -->

<!-- ```{r} -->
<!-- split_by=NULL -->
<!-- ## Join test results and dimensionality reductions -->
<!-- rdim_df <- data.frame(nhoodReducedDim(endo_milo, "UMAP")[,c(1,2)]) -->
<!-- colnames(rdim_df) <- c('X','Y') -->

<!-- n_nhoods <- length(nhoods(endo_milo)) -->
<!-- rdim_df[,"Nhood"] <- ifelse(1:nrow(rdim_df) %in% c(1:n_nhoods), c(1:n_nhoods), NA) -->
<!-- viz_df  <- left_join(rdim_df, milo_res[which(milo_res$annotation_lineage=="Endothelia"),], by="Nhood") -->
<!-- viz_df[["nhIndex"]] <- unlist(ifelse(!is.na(viz_df$Nhood), nhoodIndex(endo_milo)[viz_df$Nhood],NA)) -->
<!-- viz_df[is.na(viz_df["nhIndex"]),'nhIndex'] <- 1:ncol(endo_milo) # Add index also to single-cells -->

<!-- if (!is.null(split_by)){ -->
<!--   split_df <- data.frame(split_by=colData(endo_milo)[,split_by]) -->
<!--   split_df[,"nhIndex"] <- 1:nrow(split_df) -->
<!--   viz_df  <- left_join(viz_df, split_df, by="nhIndex") -->
<!-- } -->

<!-- filter_alpha=0.1 -->
<!-- ## Filter significant DA nhoods -->
<!-- if (!is.null(filter_alpha)) { -->
<!--   if (filter_alpha > 0) { -->
<!--     viz_df <- mutate(viz_df, logFC = ifelse(SpatialFDR > filter_alpha, NA, logFC)) -->
<!--   } -->
<!-- } -->

<!-- ## Plot -->
<!-- pt_size=1 -->
<!-- nhood_reduced_dims="UMAP" -->
<!-- components=c(1,2) -->
<!--   pl <- -->
<!--     ggplot(data = arrange(viz_df, abs(logFC)), -->
<!--            aes(X, Y)) + -->
<!--     geom_point(aes(color = ''), size = pt_size / 3, alpha = 0.5) + -->
<!--     geom_point( -->
<!--       data = . %>% filter(!is.na(SpatialFDR)), -->
<!--       aes(fill = logFC), -->
<!--       size = pt_size, -->
<!--       stroke = 0.1, -->
<!--       # colour="black", -->
<!--       shape = 21 -->
<!--     ) + -->
<!--     scale_fill_gradient2( -->
<!--       midpoint = 0, -->
<!--       high = "red", -->
<!--       low = "blue", -->
<!--       name = "log-FC" -->
<!--     ) + -->
<!--     xlab(paste(nhood_reduced_dims, components[1], sep="_")) + -->
<!--     ylab(paste(nhood_reduced_dims, components[2], sep="_")) -->

<!--   if (!is.null(split_by)) { -->
<!--     pl <- pl + facet_wrap(split_by~.) -->
<!--   } -->
<!--   if (!is.null(filter_alpha)) { -->
<!--     pl <- pl + -->
<!--       scale_color_manual(values = 'grey', label = paste("SpatialFDR >", round(filter_alpha, 2))) + -->
<!--       guides(colour = guide_legend( -->
<!--         '', -->
<!--         override.aes = list( -->
<!--           shape = 21, -->
<!--           colour = "black", -->
<!--           fill = "grey50", -->
<!--           size = pt_size, -->
<!--           alpha = 1, -->
<!--           stroke = 0.1 -->
<!--         ) -->
<!--       )) -->
<!--   } else { -->
<!--     pl <- pl + -->
<!--       scale_color_manual(values = 'grey') + -->
<!--       guides(color="none") -->
<!--   } -->

<!--   pl <- pl + -->
<!--     theme_classic(base_size = 16) + -->
<!--     theme( -->
<!--       axis.ticks = element_blank(), -->
<!--       axis.text = element_blank(), -->
<!--       plot.title = element_text(hjust = 0.5) -->
<!--     ) -->
<!-- pl -->
<!-- ``` -->

<!-- ```{r} -->
<!-- endo_milo <- scater::runPCA(endo_milo, subset_row=hvgs) -->
<!-- ``` -->



<!-- --- -->
<!-- ## Old (before I got dataset from authors) -->

<!-- Using data from [Ramachandran et al. 2019](https://www.nature.com/articles/s41586-019-1631-3#Sec1) (GEO accessiion: GSE136103).  -->

<!-- ```{r} -->
<!-- human_files <- list.files("~/Downloads/GSE136103_RAW", pattern="GSM40411.._healthy|cirrhotic", full.names = TRUE)  -->

<!-- prefixes <- str_remove(human_files, "barcodes.tsv.gz|genes.tsv.gz|matrix.mtx.gz") %>% -->
<!--   # str_remove(".+/") %>% -->
<!--   unique()  -->

<!-- sce_ls <- lapply(prefixes, function(x) read10xCounts(x, type="prefix")) -->
<!-- liver_sce <- purrr::reduce(sce_ls, cbind) -->

<!-- ## Make colData info -->
<!-- new_coldata <- colData(liver_sce) %>% -->
<!--   data.frame() %>% -->
<!--   mutate(Sample=str_remove(Sample, ".+/") %>% str_remove("_$")) %>% -->
<!--   separate(Sample, into=c("col1", "Patient", "Sort"), sep = "_", remove=FALSE) %>%  -->
<!--   mutate(Condition=str_remove(Patient, ".$")) %>% -->
<!--   select(-col1) %>% -->
<!--   mutate(Cell_id = str_c(Sample, "_",Barcode)) %>% -->
<!--   column_to_rownames("Cell_id") -->

<!-- colnames(liver_sce) <- rownames(new_coldata) -->
<!-- colData(liver_sce) <- DataFrame(new_coldata) -->

<!-- # saveRDS(liver_sce, "~/GSE136103_SingleCellExperiment.RDS") -->
<!-- ``` -->
<!-- ```{r} -->
<!-- liver_sce <- readRDS("~/GSE136103_SingleCellExperiment.RDS") -->
<!-- ``` -->

<!-- QC metrics show that the outlier cells are already filtered (following [this](https://osca.bioconductor.org/overview.html#data-processing-and-downstream-analysis)) -->

<!-- ```{r, fig.width=10,fig.height=8} -->
<!-- # is.mito <- grepl("^MT-", rownames(liver_sce)) -->
<!-- qcstats <- perCellQCMetrics(liver_sce) -->

<!-- colData(liver_sce) <- cbind(colData(liver_sce), qcstats) -->

<!-- plotColData(liver_sce, x="Sample", y = "total", other_fields = "Condition") + -->
<!--   scale_y_log10() + -->
<!--   facet_wrap(~Condition, scales = "free") + -->
<!--   ggtitle("total counts")  -->

<!-- plotColData(liver_sce, x="Sample", y = "detected", other_fields = "Condition") + -->
<!--   facet_wrap(~Condition, scales = "free") + -->
<!--   ggtitle("Detected genes")  -->

<!-- ``` -->

<!-- ### Normalization -->

<!-- ```{r} -->
<!-- lib_sf <- librarySizeFactors(liver_sce) -->
<!-- hist(log10(lib_sf), xlab="Log10[Size factor]", col='grey80') -->
<!-- ``` -->
<!-- ```{r} -->
<!-- set.seed(100) -->
<!-- liver_sce <- logNormCounts(liver_sce) -->
<!-- assayNames(liver_sce) -->
<!-- ``` -->

<!-- ### Feature selection -->

<!-- ```{r} -->
<!-- library(scran) -->
<!-- dec_liver <- modelGeneVar(liver_sce) -->

<!-- # Visualizing the fit: -->
<!-- fit_liver <- metadata(dec_liver) -->
<!-- plot(fit_liver$mean, fit_liver$var, xlab="Mean of log-expression", -->
<!--     ylab="Variance of log-expression") -->

<!-- hvgs <- getTopHVGs(dec_liver, n=3000) -->
<!-- ``` -->



<!-- ### Dim reduction -->

<!-- ```{r, fig.height=14, fig.width=14} -->
<!-- liver_sce <- scater::runPCA(liver_sce, subset_row=hvgs, ncomponents=30) -->
<!-- reducedDim(liver_sce, "PCA") <- reducedDim(liver_sce, "PCA")[,1:11] -->
<!-- plotPCA(liver_sce, colour_by="Condition", ncomponents=3) -->
<!-- ``` -->

<!-- ```{r, fig.height=8, fig.width=8} -->
<!-- liver_sce <- runUMAP(liver_sce, dimred="PCA", ncomponents=2) -->

<!-- scater::plotUMAP(liver_sce, colour_by="Condition", point_alpha=1,  point_size=0.8)  -->
<!-- scater::plotUMAP(liver_sce, colour_by="Sort", point_alpha=0.3,  point_size=0.5) -->
<!-- scater::plotUMAP(liver_sce, colour_by="Patient", point_alpha=0.3,  point_size=0.5) -->
<!-- ``` -->
<!-- ```{r, fig.height=10, fig.width=10} -->
<!-- rownames(liver_sce) <- rowData(liver_sce)$Symbol -->
<!-- scater::plotUMAP(liver_sce, colour_by="CD3D", point_alpha=0.3, point_size=0.5) -->
<!-- ``` -->



